---
title: "BDAT 608 | Module 4, Lesson 1 — SOLUTION"
author: "Instructor Reference"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r libraries}
library(tidyverse)
library(modelr)
options(na.action = na.warn)
```

## Task 1 — model_matrix() Exploration

```{r model-matrix}
df <- tribble(~y, ~x1, ~x2, 4, 2, 5, 5, 1, 6)

model_matrix(df, y ~ x1)        # intercept + x1
model_matrix(df, y ~ x1 - 1)    # x1 only, no intercept
model_matrix(df, y ~ x1 + x2)   # intercept + x1 + x2
```

## Task 2 — Categorical Predictor on sim2

```{r categorical-model}
data("sim2", package = "modelr")
mod2 <- lm(y ~ x, data = sim2)
model_matrix(sim2, y ~ x) |> head(8)
coef(mod2)
```

## Task 3 — Group Predictions

```{r group-predictions}
preds_cat <- sim2 |>
  data_grid(x) |>
  add_predictions(mod2)
print(preds_cat)
```

## Task 4 — Plot

```{r cat-plot}
ggplot(sim2, aes(x, y)) +
  geom_jitter(colour = "grey50", width = 0.1, alpha = 0.5) +
  geom_point(data = preds_cat, aes(y = pred),
             colour = "#C9A84C", size = 4, shape = 18) +
  labs(title = "sim2: OLS predictions = group means") +
  theme_minimal(base_size = 13)
```

## Reflection

The coefficient for `xb` (6.15) is **not** the mean of group b — it is the **difference** between the mean of group b and the mean of the reference group (group a). The intercept (1.15) is the mean of group a. So the mean of group b is 1.15 + 6.15 = 7.30. This is the fundamental interpretation of dummy coefficients: each one represents a deviation from the reference level, not the group mean itself.
