---
title: "BDAT 602 — Module 1, Lesson 2 Exercise"
subtitle: "CRISP-DM and Setting Up the Course Dataset"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Task 1 — Read data from CSV

Read `data/raw/health_insurance_small.csv` using `readr::read_csv()`. Call `glimpse()` on the result.

```{r task1}
# YOUR CODE HERE
# library(readr)
# df <- read_csv("data/raw/health_insurance_small.csv")
# glimpse(df)
```

**How many rows?** ___
**How many columns?** ___
**Which columns were parsed as `<lgl>` (logical)?** ___

## Task 2 — Histogram of claim amounts

Filter to rows where `claim_amount > 0`, then plot a histogram on a log10 x-axis. Title: "Claim Amounts: 10k Sample".

```{r task2}
# YOUR CODE HERE
# library(ggplot2)
# library(dplyr)
# ggplot(df |> filter(claim_amount > 0), aes(x = claim_amount)) +
#   ...
```

## Task 3 — Missing value map

Use `DataExplorer::plot_missing()` to visualise missing values. List which columns have missing data.

```{r task3}
# YOUR CODE HERE
# library(DataExplorer)
# plot_missing(df)
```

**Columns with missing values:** ___
