---
title: "Exercise: CRISP-DM and Setting Up the Course Dataset"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Task 1: Load packages and generate the dataset

```{r load-packages}
library(tidyverse)
library(skimr)
# sparklyr is loaded separately in the Spark chunk below
```

```{r task1-generate}
# TODO: source the simulator with bdat602.source_only = TRUE to prevent auto-run
options(bdat602.source_only = TRUE)
# source("courses/bdat-602/_teacher/resources/datasets/simulate_bdat602_data.R")

# TODO: call simulate_bdat602(n = 500000, seed = 602)
health_data <- NULL  # replace NULL

# TODO: print dimensions with dim() and confirm 500000 rows and 40 columns
```

## Task 2: CRISP-DM Phase 1 — Business Understanding for fraud detection

Write a short (4–6 sentence) business understanding paragraph for a fraud detection project using this dataset. Answer: What is the business problem? What is the mining objective? What success criterion would you set? Replace the placeholder below.

```{r task2-business-understanding}
# TODO: Write your Phase 1 document as R comments here.
# Business problem:
# Mining objective:
# Success criterion:
# Data source:
# Constraints:
```

## Task 3: Explore the three target variables

Use `skim()` on just the three target columns, then create at least one visualisation.

```{r task3-targets-skim}
# TODO: select churned, high_risk, claim_amount from health_data, then skim()
```

```{r task3-plot}
# TODO: create a ggplot histogram of claim_amount (hint: filter claim_amount > 0 first)
# Then describe the shape of the distribution in a comment.

# Distribution shape:
```

```{r task3-churn-rate}
# TODO: compute the churn rate (mean of churned) and the high_risk rate (mean of high_risk)
# Compare them — which target is more imbalanced?
```

## Task 4: Spark connection (eval=FALSE — do not run)

```{r task4-spark, eval=FALSE}
library(sparklyr)

# TODO: connect to local Spark
sc <- spark_connect(master = "???", version = "3.4.1")

# TODO: push health_data to Spark with copy_to(), name = "health_insurance"
health_tbl <- NULL  # replace NULL

# TODO: count records by plan_tier using dplyr syntax, then collect() the result
plan_counts <- NULL  # replace NULL

# TODO: disconnect from Spark
spark_disconnect(sc)
```
