---
title: "BDAT 602 — Module 1, Lesson 2 Solution"
subtitle: "CRISP-DM and Setting Up the Course Dataset"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Task 1 — Read data from CSV

```{r task1}
library(readr)

df <- read_csv("data/raw/health_insurance_small.csv")
glimpse(df)
```

The dataset has **10,000 rows** and **40 columns**. Logical columns include `dental_cover`, `vision_cover`, `mental_cover`, `maternity_cover`, `smoker`, `auto_pay`, `fraud_flag`, `churned`, `high_risk`, and `weekend_claim`.

## Task 2 — Histogram of claim amounts

```{r task2}
library(ggplot2)
library(dplyr)

ggplot(df |> filter(claim_amount > 0), aes(x = claim_amount)) +
  geom_histogram(bins = 60, fill = "steelblue", colour = "white") +
  scale_x_log10(labels = scales::comma) +
  labs(title = "Claim Amounts: 10k Sample",
       x = "Claim Amount (USD, log scale)", y = "Count") +
  theme_minimal()
```

The distribution is right-skewed with many small claims and a long tail of large ones. Using a log scale reveals the shape of the bulk distribution.

## Task 3 — Missing value map

```{r task3}
library(DataExplorer)

plot_missing(df)
```

Columns with missing values: `bmi` (~5%), `income` (~4%), `days_since_last_claim` (records with zero claims), `last_claim_date` (same), `weekend_claim` (same), and `complaint_notes` (~10%).
