---
title: "BDAT 608 | Module 2, Lesson 2 — SOLUTION"
author: "Instructor Reference"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r libraries}
library(tidyverse)
library(modelr)
options(na.action = na.warn)
data("sim1", package = "modelr")

model1 <- function(a, data) a[1] + data$x * a[2]
measure_distance <- function(mod, data) sqrt(mean((data$y - model1(mod, data))^2))
sim1_dist <- function(a1, a2) measure_distance(c(a1, a2), sim1)
```

## Task 1 — Random Search

```{r random-search}
set.seed(7)
models <- tibble(
  a1 = runif(500, -20, 40),
  a2 = runif(500, -5, 5)
) |>
  mutate(dist = purrr::map2_dbl(a1, a2, sim1_dist))
```

```{r top10-plot}
ggplot(sim1, aes(x, y)) +
  geom_point(size = 2, colour = "grey30") +
  geom_abline(
    aes(intercept = a1, slope = a2, colour = -dist),
    data      = filter(models, rank(dist) <= 10),
    linewidth = 0.8
  ) +
  scale_colour_viridis_c(guide = "none") +
  labs(
    title    = "Top 10 randomly sampled models (ranked by RMSE)",
    subtitle = "Brighter lines have lower RMSE."
  ) +
  theme_minimal(base_size = 13)
```

## Task 2 — RMSE Landscape

```{r rmse-landscape}
ggplot(models, aes(a1, a2, colour = dist)) +
  geom_point(size = 1.5) +
  scale_colour_viridis_c(name = "RMSE") +
  labs(
    title = "RMSE landscape over (intercept, slope)",
    x = expression(a[1] ~ "(intercept)"),
    y = expression(a[2] ~ "(slope)")
  ) +
  theme_minimal(base_size = 13)
```

The low-RMSE region (yellow) is a narrow diagonal valley — the true minimum lies at approximately a1 ≈ 4.2, a2 ≈ 2.0.

## Task 3 — optim()

```{r optim-fit}
best <- optim(par = c(0, 0), fn = measure_distance, data = sim1)
cat("Best intercept:", round(best$par[1], 4), "\n")
cat("Best slope:    ", round(best$par[2], 4), "\n")
cat("Minimum RMSE:  ", round(best$value,  4), "\n")
```

## Task 4 — Comparison

```{r comparison}
best_random <- models |> slice_min(dist, n = 1)
cat("Best random search — a1:", round(best_random$a1, 4),
    "  a2:", round(best_random$a2, 4), "  RMSE:", round(best_random$dist, 4), "\n")
cat("optim()             — a1:", round(best$par[1], 4),
    "  a2:", round(best$par[2], 4),  "  RMSE:", round(best$value, 4), "\n")
```

`optim()` consistently finds a lower RMSE than the best random sample because it uses gradient information to navigate the loss landscape efficiently, whereas random search samples blindly.

## Reflection

If `optim()` starts far from the true minimum (e.g., `par = c(100, 100)` for this problem), the Nelder-Mead algorithm may converge to a local minimum or require many more iterations. For a convex loss (like RMSE), there is only one minimum, so starting values matter mainly for speed. For non-convex losses, poor starting values can lead to a suboptimal solution. Mitigation strategies: (1) use a coarse grid or random search to identify a promising region first, then hand that to `optim()`; (2) run `optim()` from several different starting points and keep the best result.
