---
title: "BDAT 608 | Module 6, Lesson 1 — SOLUTION"
author: "Instructor Reference"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5, fig.align = "center")
```

```{r libraries}
library(tidyverse)
library(modelr)
library(nycflights13)
library(broom)
options(na.action = na.warn)
data("diamonds", package = "ggplot2")
select <- dplyr::select
filter <- dplyr::filter
```

# Case Study 1: Diamonds

## EDA

```{r diamonds-eda}
glimpse(diamonds)

ggplot(diamonds, aes(carat, price)) +
  geom_bin2d(bins = 60) + scale_fill_viridis_c() +
  labs(title = "Price vs Carat (raw scale)") + theme_minimal(base_size = 13)

ggplot(diamonds, aes(log(carat), log(price))) +
  geom_bin2d(bins = 60) + scale_fill_viridis_c() +
  labs(title = "log(price) vs log(carat) — much more linear!") + theme_minimal(base_size = 13)
```

## Log-Log Model

```{r log-log-model}
mod_d1 <- lm(log(price) ~ log(carat), data = diamonds)
cat("R²:", round(summary(mod_d1)$r.squared, 3), "\n")
coef(mod_d1)
# Slope ≈ 1.68: a 1% increase in carat → ~1.68% increase in price (power law)
```

## Adding Quality Attributes

```{r quality-model}
mod_d2 <- lm(log(price) ~ log(carat) + cut + color + clarity, data = diamonds)
cat("R² (carat only):   ", round(summary(mod_d1)$r.squared, 3), "\n")
cat("R² (full model):   ", round(summary(mod_d2)$r.squared, 3), "\n")
AIC(mod_d1, mod_d2)
```

## Residuals by Cut

```{r diamonds-residuals}
diamonds |>
  add_residuals(mod_d2, var = "resid2") |>
  ggplot(aes(log(carat), resid2)) +
  geom_bin2d(bins = 50) + scale_fill_viridis_c() +
  geom_ref_line(h = 0) +
  facet_wrap(~ cut) +
  labs(title = "Residuals by cut — after controlling for carat", y = "Residual (log scale)") +
  theme_minimal(base_size = 11)
```

# Case Study 2: Flights

## Fit Per-Carrier Models

```{r carrier-models}
carrier_models <- flights |>
  filter(!is.na(dep_delay)) |>
  group_by(carrier) |>
  nest() |>
  mutate(
    mod  = map(data, ~ lm(dep_delay ~ hour + month, data = .x)),
    rsq  = map_dbl(mod, ~ summary(.x)$r.squared),
    rmse = map_dbl(mod, ~ sqrt(mean(residuals(.x)^2)))
  ) |>
  arrange(desc(rsq))

carrier_models |> select(carrier, rsq, rmse)
```

## Extract Hour Coefficients

```{r carrier-coefs}
carrier_coefs <- carrier_models |>
  mutate(tidy = map(mod, broom::tidy)) |>
  unnest(tidy) |>
  filter(term == "hour")

ggplot(carrier_coefs,
       aes(x = reorder(carrier, estimate), y = estimate,
           ymin = estimate - 2*std.error, ymax = estimate + 2*std.error)) +
  geom_pointrange(colour = "#1B3A6B") +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "#C9A84C") +
  coord_flip() +
  labs(title = "Effect of hour on departure delay, by carrier",
       x = "Carrier", y = "Estimated slope (min per hour)") +
  theme_minimal(base_size = 12)
```

# Case Study 3: Logistic Growth

## Simulate Data

```{r logistic-data}
set.seed(5)
t_seq <- seq(0, 20, by = 0.5)
logistic_growth <- tibble(
  t = t_seq,
  y = 100 / (1 + exp(-0.5 * (t - 10))) + rnorm(length(t_seq), sd = 3)
)
ggplot(logistic_growth, aes(t, y)) +
  geom_point(colour = "#1B3A6B") + theme_minimal(base_size = 13)
```

## Fit nls()

```{r nls-fit}
mod_nls <- nls(
  y ~ K / (1 + exp(-r * (t - t0))),
  data  = logistic_growth,
  start = list(K = 100, r = 0.5, t0 = 10)
)
coef(mod_nls)
```

## Plot Fitted S-Curve

```{r nls-plot}
logistic_growth |>
  data_grid(t = seq_range(t, 60)) |>
  add_predictions(mod_nls) |>
  ggplot(aes(t, pred)) +
  geom_line(colour = "#C9A84C", linewidth = 1.2) +
  geom_point(data = logistic_growth, aes(y = y), colour = "#1B3A6B", alpha = 0.6) +
  labs(title = "Logistic growth: nls() fit",
       subtitle = paste0("K=", round(coef(mod_nls)["K"],1),
                         "  r=", round(coef(mod_nls)["r"],3),
                         "  t0=", round(coef(mod_nls)["t0"],2))) +
  theme_minimal(base_size = 13)
```

# Extension: Largest Residuals

```{r largest-residuals}
diamonds_500   <- diamonds |> slice_head(n = 500)
mod_d1_500     <- lm(log(price) ~ log(carat), data = diamonds_500)

broom::augment(mod_d1_500, data = diamonds_500) |>
  mutate(abs_resid = abs(.resid)) |>
  arrange(desc(abs_resid)) |>
  slice_head(n = 5) |>
  select(carat, cut, color, clarity, price, .fitted, .resid)
```

**Pattern:** Large positive residuals (actual >> predicted) tend to be diamonds with exceptional quality (Ideal/Premium cut, colourless D/E colour, high clarity). Large negative residuals are diamonds with poor quality that are "overpriced" relative to their carat weight.

## Reflection

All three case studies follow the same core cycle: **fit → visualise residuals → diagnose → revise**. In diamonds, residuals by cut revealed missing quality predictors. In flights, R² near zero revealed that time and month cannot explain delays — prompting us to question the variable selection. In the NLS case, providing good starting values (from domain knowledge: K = carrying capacity, t0 = inflection time) was essential because without them the iterative algorithm cannot converge. The unifying lesson: residuals tell you what your model missed — reading them is a skill more important than knowing any particular model family.
