---
title: "BDAT 608 | Module 3, Lesson 2 — SOLUTION"
author: "Instructor Reference"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5)
```

```{r libraries}
library(tidyverse)
library(modelr)
library(ggfortify)
options(na.action = na.warn)
data("sim1", package = "modelr")
sim1_mod <- lm(y ~ x, data = sim1)
```

## Task 1 — Base R Diagnostics

```{r base-diagnostics, fig.width=9, fig.height=7}
par(mfrow = c(2, 2))
plot(sim1_mod)
par(mfrow = c(1, 1))
```

All four panels look clean for `sim1`: random scatter around zero, points near the Q-Q diagonal, flat Scale-Location line, and small Cook's distances.

## Task 2 — ggfortify Diagnostics

```{r ggplot-diagnostics}
autoplot(sim1_mod, which = 1:4, colour = "#1B3A6B", smooth.colour = "#C9A84C")
```

## Task 3 — Fit Quadratic Model

```{r quadratic-model}
mod_b <- lm(y ~ x + I(x^2), data = sim1)
summary(mod_b)
```

## Task 4 — Compare Fitted Lines

```{r compare-fits}
mod_a <- sim1_mod

grid_multi <- sim1 |>
  data_grid(x) |>
  gather_predictions(mod_a, mod_b)

ggplot(sim1, aes(x, y)) +
  geom_point(colour = "grey40", size = 2) +
  geom_line(data = grid_multi, aes(y = pred, colour = model), linewidth = 1.2) +
  scale_colour_manual(values = c("#1B3A6B", "#C9A84C"),
                      labels = c("Linear", "Quadratic")) +
  labs(title = "Linear vs Quadratic model on sim1", colour = "Model") +
  theme_minimal(base_size = 13)
```

## Task 5 — Compare Residuals

```{r compare-residuals}
sim1 |>
  gather_residuals(mod_a, mod_b) |>
  ggplot(aes(x, resid, colour = model)) +
  geom_point() +
  geom_ref_line(h = 0) +
  facet_wrap(~ model, labeller = label_both) +
  scale_colour_manual(values = c("#1B3A6B", "#C9A84C")) +
  labs(title = "Residuals: linear vs quadratic", y = "Residual") +
  theme_minimal(base_size = 13)
```

## Task 6 — AIC Comparison

```{r aic}
AIC(mod_a, mod_b)
cat("Additive R²:    ", round(summary(mod_a)$r.squared, 4), "\n")
cat("Quadratic R²:   ", round(summary(mod_b)$r.squared, 4), "\n")
# The linear model (mod_a) has a lower (better) AIC despite fewer parameters.
# For sim1, the relationship is genuinely linear — the quadratic term adds complexity
# without capturing any real additional signal.
```

## Reflection

Non-normality of residuals does **not** automatically mean the model is wrong. Normality matters primarily for **inference** — t-tests, p-values, and confidence intervals rely on the normality assumption. For large samples, the Central Limit Theorem ensures these are approximately valid even with moderately non-normal residuals. However, normality does not affect the unbiasedness or consistency of the OLS estimates themselves. The more critical assumptions are **linearity of the mean** (checked by Residuals vs Fitted) and **homoscedasticity** (checked by Scale-Location). Heavy tails in residuals typically indicate that standard errors and p-values should be treated with caution, and robust regression (`rlm()`) may be more appropriate.
