---
title: "BDAT 602 — Module 2, Lesson 3 Exercise"
subtitle: "Scaling, Encoding, and the recipes Pipeline"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
source("R/simulate_bdat602_data.R")
health_small <- simulate_bdat602(n = 10000, seed = 602)
```

## Task 1 — Full preprocessing pipeline

Build a `recipes` pipeline that:
- imputes `bmi` and `income` with their medians (`step_impute_median`)
- dummy-encodes `sex` and `region` (one_hot = FALSE)
- Z-score normalises all numeric predictors (`step_normalize`)

`prep()` on `health_small`, then `bake()` to get `health_processed`. Verify:
- zero NAs in `bmi` and `income`
- `mean(health_processed$bmi)` is approximately 0

```{r task1}
# YOUR CODE HERE
# library(recipes)
# pp_rec <- recipe(claim_amount ~ ., data = health_small) |>
#   step_impute_median(bmi, income) |>
#   step_dummy(sex, region, one_hot = FALSE) |>
#   step_normalize(all_numeric_predictors())
# pp_prep       <- prep(pp_rec, training = health_small)
# health_processed <- bake(pp_prep, new_data = health_small)
# sum(is.na(health_processed$bmi))
# round(mean(health_processed$bmi), 4)
```

## Task 2 — Column count before and after step_nzv

Report how many columns `health_processed` has (from Task 1). Then add `step_nzv(all_predictors())` to the pipeline and report the new column count. Which columns (if any) were dropped?

```{r task2}
# YOUR CODE HERE
# ncol(health_processed)
# pp_rec2 <- recipe(claim_amount ~ ., data = health_small) |>
#   step_impute_median(bmi, income) |>
#   step_dummy(sex, region, one_hot = FALSE) |>
#   step_normalize(all_numeric_predictors()) |>
#   step_nzv(all_predictors())
# pp_prep2  <- prep(pp_rec2, training = health_small)
# health_p2 <- bake(pp_prep2, new_data = health_small)
# ncol(health_p2)
```
