---
title: "BDAT 608 | Module 2, Lesson 1 — SOLUTION"
author: "Instructor Reference"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r libraries}
library(tidyverse)
library(modelr)
options(na.action = na.warn)
data("sim1", package = "modelr")
```

## Task 1 — Implement Loss Functions

```{r loss-functions}
model1 <- function(a, data) {
  a[1] + data$x * a[2]
}

measure_distance <- function(mod, data) {
  resid <- data$y - model1(mod, data)
  sqrt(mean(resid^2))
}

measure_mae <- function(mod, data) {
  resid <- data$y - model1(mod, data)
  mean(abs(resid))
}
```

## Task 2 — Compare Candidates on sim1

```{r compare-candidates}
candidates <- list(c1 = c(7, 1.5), c2 = c(5, 2.0), c3 = c(4, 2.0))

results_sim1 <- tibble(
  candidate = names(candidates),
  intercept = sapply(candidates, `[`, 1),
  slope     = sapply(candidates, `[`, 2),
  RMSE      = sapply(candidates, measure_distance, data = sim1),
  MAE       = sapply(candidates, measure_mae,      data = sim1)
)
print(results_sim1)
```

**Lowest RMSE on sim1:** candidate `c3` (intercept = 4, slope = 2) has the lowest RMSE and MAE among these three. The true OLS fit is approximately intercept = 4.22, slope = 2.05, so `c3` is closest.

## Task 3 — Heavy-Tailed Noise

```{r heavy-tailed}
set.seed(42)
sim1a <- tibble(
  x = rep(1:10, each = 3),
  y = x * 1.5 + 6 + rt(length(x), df = 2)
)

results_sim1a <- tibble(
  candidate = names(candidates),
  RMSE      = sapply(candidates, measure_distance, data = sim1a),
  MAE       = sapply(candidates, measure_mae,      data = sim1a)
)
print(results_sim1a)
```

Under heavy-tailed noise the two criteria may rank candidates differently because Student-$t$(df=2) occasionally produces extreme values that inflate RMSE much more than MAE.

## Reflection

RMSE and MAE measure "closeness" differently: RMSE squares each residual (amplifying large errors), while MAE takes the absolute value (treating all magnitudes linearly). A candidate with one very large residual will have a high RMSE but a moderate MAE. Therefore the "best" candidate depends on the loss — different loss functions define different optimal parameters. The loss function is not a neutral technical choice; it encodes an assumption about what kind of error matters more.
