---
title: "BDAT 602 — Module 4, Lesson 2 Exercise"
subtitle: "Hierarchical Clustering"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
source("R/simulate_bdat602_data.R")
health_small <- simulate_bdat602(n = 10000, seed = 602)
```

## Task 1 — Prepare data and distance matrix

Take a 500-row sample of `health_small` (use `set.seed(602)` before `slice_sample()`). Select `age`, `bmi`, and `num_claims`. Impute `bmi` with its median. Scale all three variables. Compute the Euclidean distance matrix.

```{r task1}
# YOUR CODE HERE
# library(dplyr)
# set.seed(602)
# hc_sub <- health_small |>
#   slice_sample(n = 500) |>
#   select(age, bmi, num_claims) |>
#   mutate(bmi = if_else(is.na(bmi), median(bmi, na.rm=TRUE), bmi)) |>
#   na.omit()
# hc_scaled <- scale(hc_sub)
# dist_mat  <- dist(hc_scaled, method = "euclidean")
```

## Task 2 — Fit hierarchical clustering and plot dendrogram

Fit with Ward.D2 linkage. Plot the dendrogram with `labels = FALSE`, `hang = -1`.

```{r task2}
# YOUR CODE HERE
# hc_ward <- hclust(dist_mat, method = "ward.D2")
# plot(hc_ward, labels = FALSE, hang = -1,
#      main = "Dendrogram: Ward.D2 Linkage", xlab = "", ylab = "Height")
```

## Task 3 — Cut at k = 3

Use `cutree()` to cut at k = 3. Print the cluster size table.

```{r task3}
# YOUR CODE HERE
# clusters <- cutree(hc_ward, k = 3)
# table(clusters)
```
