---
title: "BDAT 602 — Module 4, Lesson 3 Exercise"
subtitle: "PCA for Dimensionality Reduction and Cluster Visualisation"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
source("R/simulate_bdat602_data.R")
health_small <- simulate_bdat602(n = 10000, seed = 602)
```

## Task 1 — Fit PCA pipeline

Build a `recipes` PCA pipeline on `age`, `bmi`, `income`, `num_claims`, `claim_amount`, and `num_chronic_conditions`. Steps: impute medians for `bmi` and `income`, normalise all numeric predictors, apply PCA with `num_comp = 3`. `prep()` and `bake()` it.

```{r task1}
# YOUR CODE HERE
# library(recipes)
# pca_vars <- c("age", "bmi", "income", "num_claims",
#               "claim_amount", "num_chronic_conditions")
# pca_rec <- recipe(~ ., data = health_small[, pca_vars]) |>
#   step_impute_median(bmi, income) |>
#   step_normalize(all_numeric_predictors()) |>
#   step_pca(all_numeric_predictors(), num_comp = 3)
# pca_prep   <- prep(pca_rec)
# pca_scores <- bake(pca_prep, new_data = NULL)
# head(pca_scores)
```

## Task 2 — Scree plot

Using `broom::tidy(pca_prep, number = ..., type = "variance")`, extract the percent variance per PC and plot a scree plot (bar chart with a line overlay).

```{r task2}
# YOUR CODE HERE
# library(broom); library(ggplot2)
# tidy(pca_prep, number = 3, type = "variance") |>
#   filter(terms == "percent variance") |>
#   ggplot(aes(x = component, y = value)) +
#   geom_col(fill = "steelblue") +
#   geom_line(group = 1, colour = "red") +
#   labs(title = "Scree Plot", x = "PC", y = "% Variance Explained") +
#   theme_minimal()
```

## Task 3 — Score plot coloured by plan_tier

Join `plan_tier` from `health_small` (using the same rows that entered PCA — omit NAs on `bmi`/`income`/`num_chronic_conditions` first). Plot PC1 vs PC2 coloured by `plan_tier`.

```{r task3}
# YOUR CODE HERE
# library(dplyr)
# plan <- health_small |>
#   select(bmi, income, num_chronic_conditions, plan_tier) |>
#   mutate(bmi    = if_else(is.na(bmi),    median(bmi,    na.rm=TRUE), bmi),
#          income = if_else(is.na(income), median(income, na.rm=TRUE), income)) |>
#   na.omit() |>
#   pull(plan_tier)
# pca_scores |>
#   mutate(plan_tier = plan) |>
#   ggplot(aes(x = PC1, y = PC2, colour = plan_tier)) +
#   geom_point(alpha = 0.2, size = 0.5) +
#   labs(title = "PCA Score Plot by Plan Tier") +
#   theme_minimal()
```
