---
title: "Exercise: What Is Data Mining? The Big Data Landscape and the KDD Process"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Setup: load packages

```{r load-packages}
library(tidyverse)
library(skimr)

# Source the dataset simulator (prevents auto-run)
options(bdat602.source_only = TRUE)
source("courses/bdat-602/_teacher/resources/datasets/simulate_bdat602_data.R")
```

## Task 1: Generate the dataset

Generate the full 500,000-row insurance dataset and inspect its structure.

```{r task1-generate}
# TODO: call simulate_bdat602(n = 500000, seed = 602) and assign to health_data
health_data <- NULL  # replace NULL with your call

# TODO: use glimpse() to inspect the dataset
```

## Task 2: First look with slice_sample

Take a random sample of 10 rows to see what a typical record looks like.

```{r task2-sample}
# TODO: use slice_sample(n = 10) on health_data and print the result
```

## Task 3: Run skim() and identify three findings

Run `skim()` on the full dataset, then write your three findings as comments below each sub-task.

```{r task3-skim}
# TODO: call skim(health_data)
```

```{r task3-findings}
# Finding 1 (numeric variable):
# e.g. "The variable ___ has mean ___ and is right-skewed because ___"
# YOUR ANSWER:

# Finding 2 (character/factor variable):
# e.g. "The variable ___ has ___ unique values; the most frequent is ___"
# YOUR ANSWER:

# Finding 3 (missingness):
# e.g. "The variable ___ has ___% missing values; this could be because ___"
# YOUR ANSWER:
```

## Task 4: Map the six mining tasks to this dataset

For each mining task, write one sentence describing what you would mine from the health insurance data. Replace "YOUR SENTENCE" with your answer.

```{r task4-mining-tasks}
mining_tasks <- tribble(
  ~task,               ~your_application,
  "Classification",    "YOUR SENTENCE",
  "Regression",        "YOUR SENTENCE",
  "Clustering",        "YOUR SENTENCE",
  "Association Rules", "YOUR SENTENCE",
  "Anomaly Detection", "YOUR SENTENCE",
  "Text Mining",       "YOUR SENTENCE"
)

print(mining_tasks)
```
