---
title: "BDAT 602 — Module 1, Lesson 1 Exercise"
subtitle: "What Is Data Mining? The Big Data Landscape and the KDD Process"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Task 1 — Generate the course dataset

Source the data generator and create `health_small` (10,000 rows). Report the dimensions and column names.

```{r task1}
# YOUR CODE HERE
# source("R/simulate_bdat602_data.R")
# health_small <- simulate_bdat602(...)
# dim(health_small)
# names(health_small)
```

## Task 2 — First-pass EDA with skimr

Run `skimr::skim()` on `health_small`. From the output, report how many numeric columns and how many character columns the dataset has.

```{r task2}
# YOUR CODE HERE
# library(skimr)
# skim(health_small)
```

**Answer:** The dataset has ___ numeric columns and ___ character columns.

## Task 3 — Churn rate

Compute the churn rate as a percentage (proportion of rows where `churned == 1`, multiplied by 100, rounded to 2 decimal places).

```{r task3}
# YOUR CODE HERE
# library(dplyr)
# health_small |>
#   summarise(churn_rate_pct = ...)
```
