---
title: "Solution: What Is Data Mining? The Big Data Landscape and the KDD Process"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Setup: load packages

```{r load-packages}
library(tidyverse)
library(skimr)
library(DataExplorer)

options(bdat602.source_only = TRUE)
source("courses/bdat-602/_teacher/resources/datasets/simulate_bdat602_data.R")
```

## Task 1: Generate the dataset

```{r task1-generate}
health_data <- simulate_bdat602(n = 500000, seed = 602)
glimpse(health_data)
```

The dataset has 500,000 rows and 40 columns. Variable types include integers (age, num_claims), doubles (income, bmi, claim_amount), characters (sex, region, plan_tier, complaint_notes), and logicals (dental_cover, fraud_flag).

## Task 2: First look with slice_sample

```{r task2-sample}
set.seed(42)
health_data |> slice_sample(n = 10)
```

Even in 10 rows we can see the richness of the dataset: variation in plan tier, income, and claim amounts, plus some rows where `complaint_notes` is NA.

## Task 3: Run skim() and identify three findings

```{r task3-skim}
skim(health_data)
```

```{r task3-findings}
# Finding 1 (numeric variable):
# claim_amount has a mean around $4,200 but a max exceeding $80,000.
# The p75 (about $6,000) is much lower than the max, indicating severe
# right skew. A log transformation will likely be needed before modelling.

# Finding 2 (character variable):
# plan_tier has 4 unique values. The most frequent is "Silver" (~35%),
# followed by "Bronze" (~30%), "Gold" (~25%), and "Platinum" (~10%).
# This mirrors realistic product distributions in insurance.

# Finding 3 (missingness):
# bmi has ~5% missing values (about 25,000 rows).
# income has ~4% missing. complaint_notes has ~10% missing.
# These are intentional and reflect real-world data collection issues.
```

## Task 4: Map the six mining tasks to this dataset

```{r task4-mining-tasks}
mining_tasks <- tribble(
  ~task,               ~application,
  "Classification",    "Predict whether a policyholder will churn (churned=1) using demographics, plan details, and claims history.",
  "Regression",        "Predict the continuous claim_amount for a new policy application given age, BMI, smoker status, and plan tier.",
  "Clustering",        "Segment policyholders into behavioural groups (e.g. low-risk young adults, high-utilisation elderly) for targeted pricing.",
  "Association Rules", "Discover which add-on cover combinations (dental, vision, mental, maternity) tend to be purchased together.",
  "Anomaly Detection", "Identify claims that deviate from expected patterns in amount, timing, or frequency to flag potential fraud.",
  "Text Mining",       "Classify complaint_notes into billing, claims, coverage, and service categories to prioritise customer service actions."
)

print(mining_tasks)
```

## Task 5: Automated EDA with DataExplorer

```{r task5-sample}
set.seed(602)
health_small <- simulate_bdat602(n = 10000, seed = 602)
```

```{r task5-missing}
plot_missing(health_small)
# The three columns with highest missingness are:
# 1. complaint_notes (~10%) — only policyholders who lodged a complaint have a value (MNAR)
# 2. bmi             (~5%)  — missingness linked to whether a GP visit was recorded (MAR)
# 3. income          (~4%)  — high earners may decline to disclose (MNAR)
# These three variables will require careful imputation decisions in Module 2.
```

```{r task5-histogram}
plot_histogram(health_small)
# claim_amount is the most right-skewed numeric variable: most claims are zero or small,
# but a long tail of large claims (>$20,000) pulls the mean far above the median.
# income also shows strong right skew. Both will need log transformation before clustering or regression.
```

```{r task5-correlation}
plot_correlation(
  health_small |> select(age, bmi, income, premium, claim_amount,
                          num_claims, num_chronic_conditions),
  title = "Correlation Matrix: Key Numeric Variables"
)
# The strongest positive correlation is between num_claims and claim_amount (~0.6-0.7):
# policyholders who file more claims naturally accumulate higher total claim costs.
# premium and coverage_amount are also positively correlated (higher-tier plans charge more
# and pay out more). num_chronic_conditions correlates moderately with num_claims,
# confirming that higher medical complexity drives utilisation.
```
