---
title: "BDAT 602 — Module 5, Lesson 2 Exercise"
subtitle: "Handling Class Imbalance"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
source("R/simulate_bdat602_data.R")
health_small <- simulate_bdat602(n = 10000, seed = 602)
set.seed(602)
idx   <- sample(nrow(health_small), 0.75 * nrow(health_small))
train <- health_small[idx, ]
test  <- health_small[-idx, ]
```

## Task 1 — Baseline random forest (no correction)

Report the fraud rate in `health_small`. Fit a random forest (ntree = 200) to predict `fraud_flag` using `age`, `income`, `num_claims`, `claim_amount`, and `plan_tier`. Predict on the test set. Report recall (true positive rate for fraud).

```{r task1}
# YOUR CODE HERE
# cat("Fraud rate:", round(mean(health_small$fraud_flag)*100, 2), "%\n")
# library(randomForest)
# rf_base <- randomForest(
#   as.factor(fraud_flag) ~ age + income + num_claims + claim_amount + plan_tier,
#   data = train |> filter(!is.na(income)), ntree = 200
# )
# pred <- predict(rf_base, test |> filter(!is.na(income)))
# actual <- test$fraud_flag[!is.na(test$income)]
# cm <- table(Predicted = pred, Actual = actual)
# recall <- cm["1","1"] / sum(cm[,"1"])
# cat("Recall (no correction):", round(recall, 3), "\n")
```

## Task 2 — Class-weighted random forest

Refit with `classwt = c("0" = 1, "1" = 30)`. Report both recall and precision on the test set.

```{r task2}
# YOUR CODE HERE
# rf_wtd <- randomForest(
#   as.factor(fraud_flag) ~ age + income + num_claims + claim_amount + plan_tier,
#   data = train |> filter(!is.na(income)),
#   ntree = 200, classwt = c("0" = 1, "1" = 30)
# )
# pred_w  <- predict(rf_wtd, test |> filter(!is.na(income)))
# cm_w    <- table(Predicted = pred_w, Actual = actual)
# recall_w  <- cm_w["1","1"] / sum(cm_w[,"1"])
# precis_w  <- cm_w["1","1"] / sum(cm_w["1",])
# cat("Recall (weighted): ", round(recall_w, 3), "\n")
# cat("Precision (weighted):", round(precis_w, 3), "\n")
```

## Task 3 — Threshold adjustment

Using the class-weighted model, get probability predictions. Adjust the threshold to 0.15. Report recall and precision at this threshold.

```{r task3}
# YOUR CODE HERE
# prob <- predict(rf_wtd, test |> filter(!is.na(income)), type = "prob")[,"1"]
# threshold <- 0.15
# pred_thr  <- as.integer(prob >= threshold)
# cm_thr    <- table(Predicted = pred_thr, Actual = actual)
# print(cm_thr)
# recall_thr  <- cm_thr["1","1"] / sum(cm_thr[,"1"])
# precis_thr  <- cm_thr["1","1"] / sum(cm_thr["1",])
# cat("Recall at 0.15: ", round(recall_thr, 3), "\n")
# cat("Precision:      ", round(precis_thr, 3), "\n")
```
