---
title: "BDAT 602 — Module 5, Lesson 2 Solution"
subtitle: "Handling Class Imbalance"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
source("R/simulate_bdat602_data.R")
health_small <- simulate_bdat602(n = 10000, seed = 602)
set.seed(602)
idx   <- sample(nrow(health_small), 0.75 * nrow(health_small))
train <- health_small[idx, ]
test  <- health_small[-idx, ]
```

## Task 1 — Baseline random forest (no correction)

```{r task1}
cat("Fraud rate:", round(mean(health_small$fraud_flag) * 100, 2), "%\n")

library(randomForest)

rf_base <- randomForest(
  as.factor(fraud_flag) ~ age + income + num_claims +
    claim_amount + plan_tier,
  data  = train |> dplyr::filter(!is.na(income)),
  ntree = 200
)

pred   <- predict(rf_base, test |> dplyr::filter(!is.na(income)))
actual <- test$fraud_flag[!is.na(test$income)]
cm     <- table(Predicted = pred, Actual = actual)
print(cm)
recall <- cm["1", "1"] / sum(cm[, "1"])
cat("Recall (no correction):", round(recall, 3), "\n")
```

Recall is very low (often < 5%) — the model mostly predicts "no fraud" because the class is rare.

## Task 2 — Class-weighted random forest

```{r task2}
rf_wtd <- randomForest(
  as.factor(fraud_flag) ~ age + income + num_claims +
    claim_amount + plan_tier,
  data    = train |> dplyr::filter(!is.na(income)),
  ntree   = 200,
  classwt = c("0" = 1, "1" = 30)
)

pred_w  <- predict(rf_wtd, test |> dplyr::filter(!is.na(income)))
cm_w    <- table(Predicted = pred_w, Actual = actual)
print(cm_w)

recall_w <- cm_w["1", "1"] / sum(cm_w[, "1"])
precis_w <- cm_w["1", "1"] / sum(cm_w["1", ])

cat("Recall (weighted):   ", round(recall_w, 3), "\n")
cat("Precision (weighted):", round(precis_w, 3), "\n")
```

With class weights, recall increases substantially but precision falls — the model now flags more true frauds but also more false alarms.

## Task 3 — Threshold adjustment

```{r task3}
prob <- predict(rf_wtd,
                test |> dplyr::filter(!is.na(income)),
                type = "prob")[, "1"]

threshold <- 0.15
pred_thr  <- as.integer(prob >= threshold)
cm_thr    <- table(Predicted = pred_thr, Actual = actual)
print(cm_thr)

recall_thr <- cm_thr["1", "1"] / sum(cm_thr[, "1"])
precis_thr <- cm_thr["1", "1"] / sum(cm_thr["1", ])

cat("Recall at 0.15:  ", round(recall_thr, 3), "\n")
cat("Precision:       ", round(precis_thr, 3), "\n")
```

Lowering the threshold increases recall further but reduces precision. The operating threshold should be set based on the business cost ratio of false negatives (missed fraud) vs false positives (wasted investigation).
