---
title: "Solution: What Is a Statistical Model?"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r load-packages}
library(tidyverse)
library(modelr)
```

## Task 1: Explore the diamonds dataset

```{r task1}
data("diamonds", package = "ggplot2")
glimpse(diamonds)
summary(diamonds)
```

## Task 2: Scatter plot of price vs carat coloured by cut

```{r task2}
ggplot(diamonds, aes(x = carat, y = price, colour = cut)) +
  geom_point(alpha = 0.3) +
  labs(
    title   = "Diamond price vs carat weight, coloured by cut",
    x       = "Carat",
    y       = "Price (USD)",
    colour  = "Cut"
  ) +
  theme_minimal()
```

## Task 3: Fit a simple linear model

```{r task3}
mod1 <- lm(price ~ carat, data = diamonds)
summary(mod1)
```

**Interpretation.** The fitted model is:

$$\widehat{\text{price}} = -2{,}256 + 7{,}756 \times \text{carat}$$

Every additional carat is associated with an increase of about \$7,756 in price on average. The intercept ($-\$2{,}256$) is the predicted price for a 0-carat diamond — which is meaningless, but mathematically consistent. $R^2 \approx 0.85$ means that carat alone explains about 85% of the variance in diamond price.

## Task 4: Add residuals and inspect the pattern

```{r task4}
diamonds_resid <- diamonds |>
  add_residuals(mod1)

ggplot(diamonds_resid, aes(x = carat, y = resid)) +
  geom_point(alpha = 0.2, colour = "steelblue") +
  geom_ref_line(h = 0, colour = "darkorange") +
  labs(
    title   = "Residuals vs carat — linear model on diamonds",
    x       = "Carat",
    y       = "Residual (actual - predicted price, USD)"
  ) +
  theme_minimal()

# Interpretation: The residuals fan outward as carat increases — the spread
# of residuals is much larger for heavy diamonds than for light ones.
# This is called heteroscedasticity (non-constant variance).
# It means the linear model is systematically wrong: it underestimates
# variability for large diamonds. A log transformation of both variables
# is indicated to stabilise the variance.
```
