---
title: "BDAT 608 | Module 1, Lesson 1 — SOLUTION"
author: "Instructor Reference"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Prerequisites

```{r libraries}
library(tidyverse)
library(modelr)
```

---

## Task 1 — Explore sim1

```{r explore-sim1}
data("sim1", package = "modelr")
glimpse(sim1)
# 30 rows, 2 columns: x (integer), y (double)
```

**Answer:** `sim1` has **30 rows** and **2 columns** (`x` and `y`).

---

## Task 2 — Scatter Plot

```{r scatter-plot}
ggplot(sim1, aes(x, y)) +
  geom_point(size = 3, colour = "#1B3A6B") +
  labs(title = "sim1: My first look") +
  theme_minimal()
```

The plot shows a clear upward linear trend: as `x` increases from 1 to 10, `y` increases roughly proportionally, with random scatter around the line.

---

## Reflection

"Zero residuals" on the training data means the model has **memorised** the noise in addition to the signal. Such a model will fail on new data because noise does not repeat. The goal of modelling is to capture the signal — the repeatable pattern — not the noise. A model with zero training residuals is overfit: it is too tightly tuned to this particular sample.
