---
title: "BDAT 608 | Module 4, Lesson 3 — SOLUTION"
author: "Instructor Reference"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 5)
```

```{r libraries}
library(tidyverse)
library(modelr)
library(splines)
library(caret)
library(mice)
options(na.action = na.warn)
data("sim1", package = "modelr")
sim1_base <- sim1

set.seed(1)
sim5 <- tibble(
  x = seq(0, 3.5 * pi, length = 50),
  y = 4 * sin(x) + rnorm(50)
)
```

## Task 1 — Fit Spline Models

```{r spline-models}
sp_mods <- list(
  df1 = lm(y ~ ns(x, 1), data = sim5),
  df3 = lm(y ~ ns(x, 3), data = sim5),
  df5 = lm(y ~ ns(x, 5), data = sim5)
)

grid5 <- sim5 |>
  data_grid(x = seq_range(x, n = 60, expand = 0.1)) |>
  gather_predictions(!!!sp_mods, .pred = "y")

ggplot(sim5, aes(x, y)) +
  geom_point(colour = "grey40") +
  geom_line(data = grid5, colour = "#C9A84C", linewidth = 1) +
  facet_wrap(~ model, labeller = label_both) +
  labs(title = "Natural splines: df = 1, 3, 5") +
  theme_minimal(base_size = 12)
```

## Task 2 — 10-fold Cross-Validation

```{r cv-splines}
set.seed(42)
cv_rmse <- sapply(1:8, function(df) {
  folds <- createFolds(sim5$y, k = 10)
  mean(sapply(folds, function(idx) {
    train <- sim5[-idx, ]; test <- sim5[idx, ]
    mod  <- lm(y ~ ns(x, df), data = train)
    pred <- predict(mod, newdata = test)
    sqrt(mean((test$y - pred)^2))
  }))
})

optimal_df <- which.min(cv_rmse)
tibble(df = 1:8, cv_rmse) |>
  ggplot(aes(df, cv_rmse)) +
  geom_line(colour = "#1B3A6B", linewidth = 1) +
  geom_point(colour = "#C9A84C", size = 3) +
  geom_vline(xintercept = optimal_df, linetype = "dashed", colour = "darkred") +
  labs(
    title    = "10-fold CV RMSE for ns(x, df)",
    subtitle = paste0("Optimal df = ", optimal_df),
    x = "Degrees of freedom", y = "CV RMSE"
  ) +
  theme_minimal(base_size = 13)
```

## Task 3 — na.exclude vs na.omit

```{r na-handling}
df_na <- tribble(~x, ~y, 1, 2.2, 2, NA, 3, 3.5, 4, 8.3, NA, 10)

mod_omit   <- lm(y ~ x, data = df_na)                              # default na.omit
mod_excl   <- lm(y ~ x, data = df_na, na.action = na.exclude)

cat("nobs (na.omit):   ", nobs(mod_omit), "\n")
cat("nobs (na.exclude):", nobs(mod_excl), "\n")

cat("\nresiduals (na.omit):   "); print(residuals(mod_omit))
cat("residuals (na.exclude):"); print(residuals(mod_excl))
# na.exclude preserves positions 2 and 5 as NA in the residual vector.
```

## Task 4 — Multiple Imputation

```{r mice-imputation}
set.seed(10)
df_miss <- sim1_base |>
  mutate(y = ifelse(runif(n()) < 0.1, NA, y))

imp     <- mice(df_miss, m = 5, method = "pmm", printFlag = FALSE)
fit_imp <- with(imp, lm(y ~ x))
pooled  <- pool(fit_imp) |> summary()
print(pooled)

cc_mod  <- lm(y ~ x, data = na.omit(df_miss))
cat("\nComplete-case slope:  ", round(coef(cc_mod)[2], 4), "\n")
cat("Multiple imputation slope:", round(pooled$estimate[2], 4), "\n")
```

## Reflection

**Few folds (e.g., 3-fold):** Each training set uses only 2/3 of the data, so the model is trained on fewer observations than the full dataset. The CV estimate will be slightly **pessimistic** (high bias) because models trained on less data tend to perform slightly worse. However, 3-fold CV is computationally cheap and has **low variance** between runs.

**Many folds (LOOCV = leave-one-out):** Each training set uses n-1 observations — almost the full dataset. The bias is very small. But the n CV estimates are highly correlated (each training set differs by just one observation), giving the CV estimate **high variance**. LOOCV is also computationally expensive for large n.

**Practical choice:** 10-fold CV is the industry standard because it balances low-enough bias and manageable variance at reasonable computational cost.
