---
title: "BDAT 608 | Module 3, Lesson 1 — SOLUTION"
author: "Instructor Reference"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r libraries}
library(tidyverse)
library(modelr)
options(na.action = na.warn)
data("sim1", package = "modelr")
sim1_mod <- lm(y ~ x, data = sim1)
```

## Task 1 — Prediction Grid

```{r prediction-grid}
grid <- sim1 |>
  data_grid(x) |>
  add_predictions(sim1_mod)
print(grid)
```

## Task 2 — Plot Data and Fitted Line

```{r fitted-line-plot}
ggplot(sim1, aes(x)) +
  geom_point(aes(y = y), colour = "#1B3A6B", size = 2.5) +
  geom_line(aes(y = pred), data = grid, colour = "#C9A84C", linewidth = 1.3) +
  labs(
    title   = "sim1: Data and Fitted Line",
    y       = expression(y ~ "/" ~ hat(y)),
    caption = "Gold line = OLS fitted values"
  ) +
  theme_minimal(base_size = 13)
```

## Task 3 — Residual Distribution

```{r residual-distribution}
sim1 <- sim1 |> add_residuals(sim1_mod)

ggplot(sim1, aes(resid)) +
  geom_freqpoly(binwidth = 0.5, colour = "#1B3A6B", linewidth = 1) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "#C9A84C") +
  labs(title = "Distribution of residuals", x = "Residual (y − ŷ)") +
  theme_minimal(base_size = 13)
```

The residuals are approximately symmetric around zero, with no obvious skew — consistent with the normality assumption.

## Task 4 — Residuals vs x

```{r residuals-vs-x}
ggplot(sim1, aes(x, resid)) +
  geom_ref_line(h = 0, colour = "#C9A84C") +
  geom_point(colour = "#1B3A6B", size = 2.5) +
  labs(title = "Residuals vs x", y = "Residual (y − ŷ)") +
  theme_minimal(base_size = 13)
# No visible pattern — random scatter around zero. The linear model is correctly specified.
```

## Reflection

A curved band in the residuals-vs-x plot (e.g., a U-shape) would indicate that the linear mean is misspecified: the true relationship between x and y is non-linear. The model is systematically over-predicting in the middle of the x range and under-predicting at the extremes (or vice versa). The next step would be to add a non-linear term to the model — for example, a quadratic term `I(x^2)` or a natural spline `ns(x, df=3)` — and recheck the residuals.
