---
title: "BDAT 602 — Module 6, Lesson 2 Exercise"
subtitle: "Topic Modelling with LDA"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
source("R/simulate_bdat602_data.R")
health_small <- simulate_bdat602(n = 10000, seed = 602)
```

## Task 1 — Build the DTM

Repeat the tokenisation pipeline from Lesson 1 (filter non-NA complaints, tokenise, remove stop words, filter nchar > 2, cast to DTM). Report the DTM dimensions.

```{r task1}
# YOUR CODE HERE
# library(tidytext); library(dplyr)
# data("stop_words")
# complaints <- health_small |>
#   filter(!is.na(complaint_notes)) |>
#   mutate(doc_id = row_number())
# dtm <- complaints |>
#   unnest_tokens(word, complaint_notes) |>
#   anti_join(stop_words, by = "word") |>
#   filter(nchar(word) > 2) |>
#   count(doc_id, word) |>
#   cast_dtm(document = doc_id, term = word, value = n)
# dim(dtm)
```

## Task 2 — Fit LDA and plot top words

Fit `LDA(dtm, k = 4, control = list(seed = 602))`. Extract the beta matrix with `tidy(..., matrix = "beta")`. Plot the top 6 words per topic using faceted bar charts.

```{r task2}
# YOUR CODE HERE
# library(topicmodels)
# set.seed(602)
# lda_fit <- LDA(dtm, k = 4, control = list(seed = 602))
# beta_mat <- tidy(lda_fit, matrix = "beta")
# beta_mat |>
#   group_by(topic) |>
#   slice_max(beta, n = 6) |>
#   ggplot(aes(x = reorder(term, beta), y = beta)) +
#   geom_col(fill = "steelblue") +
#   facet_wrap(~ topic, scales = "free_y") +
#   coord_flip() +
#   labs(title = "Top 6 Words per LDA Topic", x = NULL) +
#   theme_minimal()
```

## Task 3 — Topic prevalence

Extract the gamma matrix. Compute average gamma per topic. Which topic is most prevalent in the corpus?

```{r task3}
# YOUR CODE HERE
# gamma_mat <- tidy(lda_fit, matrix = "gamma")
# gamma_mat |>
#   group_by(topic) |>
#   summarise(avg_gamma = round(mean(gamma), 3)) |>
#   arrange(desc(avg_gamma))
```

**Most prevalent topic:** ___
