---
title: "BDAT 602 — Module 6, Lesson 2 Solution"
subtitle: "Topic Modelling with LDA"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
source("R/simulate_bdat602_data.R")
health_small <- simulate_bdat602(n = 10000, seed = 602)
```

## Task 1 — Build the DTM

```{r task1}
library(tidytext)
library(dplyr)

data("stop_words")

complaints <- health_small |>
  filter(!is.na(complaint_notes)) |>
  mutate(doc_id = row_number())

dtm <- complaints |>
  unnest_tokens(word, complaint_notes) |>
  anti_join(stop_words, by = "word") |>
  filter(nchar(word) > 2) |>
  count(doc_id, word) |>
  cast_dtm(document = doc_id, term = word, value = n)

dim(dtm)
```

## Task 2 — Fit LDA and plot top words

```{r task2}
library(topicmodels)
library(ggplot2)

set.seed(602)
lda_fit  <- LDA(dtm, k = 4, control = list(seed = 602))
beta_mat <- tidy(lda_fit, matrix = "beta")

beta_mat |>
  group_by(topic) |>
  slice_max(beta, n = 6) |>
  ungroup() |>
  ggplot(aes(x = reorder(term, beta), y = beta)) +
  geom_col(fill = "steelblue") +
  facet_wrap(~ topic, scales = "free_y") +
  coord_flip() +
  labs(title = "LDA: Top 6 Words per Topic (k=4)",
       x = NULL, y = "Beta") +
  theme_minimal()
```

With 4 topics and the templated complaint notes, the topics should map approximately to: billing/premium, claims processing, coverage/services, and positive feedback.

## Task 3 — Topic prevalence

```{r task3}
gamma_mat <- tidy(lda_fit, matrix = "gamma")

gamma_mat |>
  group_by(topic) |>
  summarise(avg_gamma = round(mean(gamma), 3)) |>
  arrange(desc(avg_gamma))
```

The most prevalent topic is likely the billing/claims topic — consistent with these being the most common complaint categories in the simulation templates.
