---
title: "BDAT 602 — Module 5, Lesson 1 Exercise"
subtitle: "Decision Trees and Random Forests"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
source("R/simulate_bdat602_data.R")
health_small <- simulate_bdat602(n = 10000, seed = 602)
set.seed(602)
idx   <- sample(nrow(health_small), 0.75 * nrow(health_small))
train <- health_small[idx, ]
test  <- health_small[-idx, ]
```

## Task 1 — Fit and plot a decision tree

Predict `high_risk` using `age`, `bmi`, `smoker`, `num_chronic_conditions`, and `plan_tier`. Set `maxdepth = 4`. Plot with `rpart.plot()`.

```{r task1}
# YOUR CODE HERE
# library(rpart); library(rpart.plot)
# tree_fit <- rpart(
#   high_risk ~ age + bmi + smoker + num_chronic_conditions + plan_tier,
#   data = train, method = "class",
#   control = rpart.control(maxdepth = 4)
# )
# rpart.plot(tree_fit, type = 4, extra = 104,
#            main = "Decision Tree: High-Risk Prediction")
```

## Task 2 — Confusion matrix and accuracy

Predict classes on `test`. Build the confusion matrix with `table()`. Report accuracy.

```{r task2}
# YOUR CODE HERE
# pred <- predict(tree_fit, test, type = "class")
# conf_mat <- table(Predicted = pred, Actual = test$high_risk)
# print(conf_mat)
# accuracy <- sum(diag(conf_mat)) / sum(conf_mat)
# cat("Accuracy:", round(accuracy, 3), "\n")
```

## Task 3 — Random forest and variable importance

Fit a random forest with `ntree = 200`, `mtry = 2`, `importance = TRUE`. Plot variable importance.

```{r task3}
# YOUR CODE HERE
# library(randomForest)
# set.seed(602)
# rf_fit <- randomForest(
#   as.factor(high_risk) ~ age + bmi + smoker +
#     num_chronic_conditions + plan_tier,
#   data = train, ntree = 200, mtry = 2, importance = TRUE
# )
# print(rf_fit)
# varImpPlot(rf_fit, main = "RF Variable Importance: High-Risk")
```
