---
title: "BDAT 602 — Module 5, Lesson 1 Solution"
subtitle: "Decision Trees and Random Forests"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
source("R/simulate_bdat602_data.R")
health_small <- simulate_bdat602(n = 10000, seed = 602)
set.seed(602)
idx   <- sample(nrow(health_small), 0.75 * nrow(health_small))
train <- health_small[idx, ]
test  <- health_small[-idx, ]
```

## Task 1 — Fit and plot a decision tree

```{r task1}
library(rpart)
library(rpart.plot)

tree_fit <- rpart(
  high_risk ~ age + bmi + smoker + num_chronic_conditions + plan_tier,
  data    = train,
  method  = "class",
  control = rpart.control(maxdepth = 4)
)

rpart.plot(tree_fit, type = 4, extra = 104,
           main = "Decision Tree: High-Risk Prediction")
```

The top splits should be on `age` and `num_chronic_conditions` — directly reflecting the rule-based definition of `high_risk` in the simulation.

## Task 2 — Confusion matrix and accuracy

```{r task2}
pred     <- predict(tree_fit, test, type = "class")
conf_mat <- table(Predicted = pred, Actual = test$high_risk)
print(conf_mat)

accuracy <- sum(diag(conf_mat)) / sum(conf_mat)
cat("Accuracy:", round(accuracy, 3), "\n")
```

Accuracy is typically > 90% — but this is partly a consequence of the explicit rule-based definition of `high_risk`. High accuracy does not imply the model is useful if the majority class dominates.

## Task 3 — Random forest and variable importance

```{r task3}
library(randomForest)

set.seed(602)
rf_fit <- randomForest(
  as.factor(high_risk) ~ age + bmi + smoker +
    num_chronic_conditions + plan_tier,
  data       = train,
  ntree      = 200,
  mtry       = 2,
  importance = TRUE
)

print(rf_fit)
varImpPlot(rf_fit, main = "RF Variable Importance: High-Risk")
```

`num_chronic_conditions` and `age` should rank as the most important variables (MeanDecreaseAccuracy and MeanDecreaseGini), consistent with the simulation definition.
