---
title: "BDAT 602 — Module 3, Lesson 1 Exercise"
subtitle: "Market Basket Analysis and the Apriori Algorithm"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Task 1 — Explore the Groceries dataset

Load `Groceries` from `arules`. Call `summary()`. Report: how many transactions? How many items?

```{r task1}
# YOUR CODE HERE
# library(arules)
# data("Groceries")
# summary(Groceries)
```

**Transactions:** ___
**Items:** ___

## Task 2 — Run Apriori

Run `apriori()` on `Groceries` with `supp = 0.01`, `conf = 0.25`, `minlen = 2`. Store the result in `rules`. Print the number of rules generated.

```{r task2}
# YOUR CODE HERE
# rules <- apriori(Groceries, parameter = list(supp = 0.01, conf = 0.25, minlen = 2))
# length(rules)
```

**Number of rules generated:** ___

## Task 3 — Top rules by lift

Print the top 5 rules sorted by lift using `inspect(head(sort(rules, by = "lift"), 5))`.

```{r task3}
# YOUR CODE HERE
```
