---
title: "BDAT 602 — Module 3, Lesson 1 Solution"
subtitle: "Market Basket Analysis and the Apriori Algorithm"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Task 1 — Explore the Groceries dataset

```{r task1}
library(arules)
data("Groceries")
summary(Groceries)
```

The `Groceries` dataset has **9,835 transactions** and **169 items**. The most frequent item is "whole milk" (appearing in ~25.5% of transactions).

## Task 2 — Run Apriori

```{r task2}
rules <- apriori(
  Groceries,
  parameter = list(supp = 0.01, conf = 0.25, minlen = 2)
)
length(rules)
```

With these thresholds, Apriori generates several hundred rules.

## Task 3 — Top rules by lift

```{r task3}
inspect(head(sort(rules, by = "lift"), 5))
```

The top rules by lift typically involve specific combinations of dairy and baking items (e.g. butter + whipped cream → whole milk), with lift values of 3–4, meaning these items co-occur 3–4× more often than chance.
