---
title: "BDAT 608 | Module 2, Lesson 3 — SOLUTION"
author: "Instructor Reference"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r libraries}
library(tidyverse)
library(modelr)
options(na.action = na.warn)
data("sim1", package = "modelr")
```

## Task 1 — Fit the Model

```{r fit-model}
sim1_mod <- lm(y ~ x, data = sim1)
coef(sim1_mod)
summary(sim1_mod)
```

## Task 2 — Interpret R²

```{r r-squared}
r2 <- summary(sim1_mod)$r.squared
cat("R-squared:", round(r2, 4), "\n")
# Interpretation: the linear model explains about 86% of the variance in y
# using x as the sole predictor.
```

## Task 3 — Compare with optim()

```{r compare-optim}
model1 <- function(a, data) a[1] + data$x * a[2]
measure_distance <- function(mod, data) sqrt(mean((data$y - model1(mod, data))^2))

best_optim <- optim(c(0, 0), measure_distance, data = sim1)

comparison <- tibble(
  method    = c("lm()", "optim()"),
  intercept = c(coef(sim1_mod)[1], best_optim$par[1]),
  slope     = c(coef(sim1_mod)[2], best_optim$par[2])
)
print(comparison)
```

The two methods give virtually identical coefficients. Any tiny difference is due to the finite convergence tolerance of the Nelder-Mead algorithm.

## Task 4 — Confidence Interval

```{r confint}
confint(sim1_mod, level = 0.95)
```

**Interpretation:** We are 95% confident that the true slope lies between approximately 1.86 and 2.24. Because the entire interval is positive, we can conclude that `x` has a statistically significant positive association with `y`.

## Reflection

`lm()` is preferred because it provides the full statistical inference table: standard errors, t-statistics, p-values, confidence intervals, and R². `optim()` returns only the parameter estimates and the minimum loss value — it tells you *where* the minimum is but not *how reliable* the estimates are. In practice, you need standard errors to test hypotheses about the coefficients and to build confidence intervals. `lm()` computes all of this analytically (via the QR decomposition) in a single call.
