---
title: "BDAT 602 — Module 3, Lesson 2 Exercise"
subtitle: "Mining Rules from the Health Insurance Dataset"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Task 1 — Build the transactions object

Generate `health_small` (n = 10000, seed = 602). Select only the four add-on cover columns (`dental_cover`, `vision_cover`, `mental_cover`, `maternity_cover`), `smoker`, and `high_risk`. Convert to a `transactions` object and call `summary()`.

```{r task1}
# YOUR CODE HERE
# source("R/simulate_bdat602_data.R")
# library(arules); library(dplyr)
# health_small <- simulate_bdat602(n = 10000, seed = 602)
# bin_cols <- c("dental_cover", "vision_cover", "mental_cover",
#               "maternity_cover", "smoker", "high_risk")
# health_bin <- health_small |>
#   select(all_of(bin_cols)) |>
#   mutate(across(everything(), as.logical))
# trans <- as(health_bin, "transactions")
# summary(trans)
```

## Task 2 — Run Apriori

Run `apriori()` with `supp = 0.05`, `conf = 0.45`, `minlen = 2`. Store in `rules_ins` and print the count.

```{r task2}
# YOUR CODE HERE
```

**Number of rules:** ___

## Task 3 — Dental cover on the RHS

Filter for rules where `dental_cover` is on the right-hand side. Print the top 3 by confidence using `inspect()`.

```{r task3}
# YOUR CODE HERE
# dental_rules <- subset(rules_ins, rhs %pin% "dental_cover")
# inspect(head(sort(dental_rules, by = "confidence"), 3))
```
