---
title: "BDAT 602 — Module 3, Lesson 2 Solution"
subtitle: "Mining Rules from the Health Insurance Dataset"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Task 1 — Build the transactions object

```{r task1}
source("R/simulate_bdat602_data.R")
library(arules)
library(dplyr)

health_small <- simulate_bdat602(n = 10000, seed = 602)

bin_cols <- c("dental_cover", "vision_cover", "mental_cover",
              "maternity_cover", "smoker", "high_risk")

health_bin <- health_small |>
  select(all_of(bin_cols)) |>
  mutate(across(everything(), as.logical))

trans <- as(health_bin, "transactions")
summary(trans)
```

The transaction summary shows item frequencies — dental cover and vision cover are among the most common (linked to Gold/Platinum tier prevalence).

## Task 2 — Run Apriori

```{r task2}
rules_ins <- apriori(
  trans,
  parameter = list(supp = 0.05, conf = 0.45, minlen = 2)
)
cat("Rules generated:", length(rules_ins), "\n")
```

## Task 3 — Dental cover on the RHS

```{r task3}
dental_rules <- subset(rules_ins, rhs %pin% "dental_cover")
inspect(head(sort(dental_rules, by = "confidence"), 3))
```

The highest-confidence dental cover rules typically involve co-purchase of vision or mental health cover, reflecting that higher-tier policyholders tend to buy the full add-on bundle.
