---
title: "BDAT 602 — Module 4, Lesson 1 Exercise"
subtitle: "k-Means Clustering"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
source("R/simulate_bdat602_data.R")
health_small <- simulate_bdat602(n = 10000, seed = 602)
```

## Task 1 — Prepare scaled clustering matrix

From `health_small`, select `age`, `bmi`, `income`, `num_claims`, and `claim_amount`. Impute `bmi` and `income` with their medians, then Z-score scale all five columns using `scale()`. Remove any rows still containing NAs. Store the result as a numeric matrix `hc_mat`.

```{r task1}
# YOUR CODE HERE
# library(dplyr)
# hc_df <- health_small |>
#   select(age, bmi, income, num_claims, claim_amount) |>
#   mutate(bmi    = if_else(is.na(bmi),    median(bmi,    na.rm=TRUE), bmi),
#          income = if_else(is.na(income), median(income, na.rm=TRUE), income)) |>
#   na.omit()
# hc_mat <- scale(hc_df)
# dim(hc_mat)
```

## Task 2 — Fit k-Means with k = 4

Run `kmeans()` with `centers = 4` and `nstart = 25`. Print the cluster sizes and total WCSS.

```{r task2}
# YOUR CODE HERE
# set.seed(602)
# km4 <- kmeans(hc_mat, centers = 4, nstart = 25)
# km4$size
# km4$tot.withinss
```

## Task 3 — Cluster profile

Join cluster assignments back to the rows of `health_small` used in Task 1 (same `na.omit()` rows). Compute per-cluster average `claim_amount` and fraud rate (`mean(fraud_flag) * 100`).

```{r task3}
# YOUR CODE HERE
# library(dplyr)
# hc_labeled <- health_small |>
#   select(age, bmi, income, num_claims, claim_amount, fraud_flag, plan_tier) |>
#   mutate(bmi    = if_else(is.na(bmi),    median(bmi,    na.rm=TRUE), bmi),
#          income = if_else(is.na(income), median(income, na.rm=TRUE), income)) |>
#   na.omit() |>
#   mutate(cluster = as.factor(km4$cluster))
# hc_labeled |>
#   group_by(cluster) |>
#   summarise(avg_claim   = round(mean(claim_amount), 0),
#             fraud_pct   = round(mean(fraud_flag) * 100, 2),
#             n           = n())
```
