---
title: "BDAT 602 — Module 2, Lesson 2 Exercise"
subtitle: "Outlier Detection and Treatment"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
source("R/simulate_bdat602_data.R")
health_small <- simulate_bdat602(n = 10000, seed = 602)
```

## Task 1 — Flag BMI outliers (IQR rule)

Compute Q1, Q3, and IQR for `bmi`. Add a `bmi_outlier` column (TRUE if below Q1 – 1.5×IQR or above Q3 + 1.5×IQR). Count how many records are flagged.

```{r task1}
# YOUR CODE HERE
# library(dplyr)
# q1  <- quantile(health_small$bmi, 0.25, na.rm = TRUE)
# ...
```

**How many BMI outliers?** ___

## Task 2 — Winsorise income

Cap `income` at the 1st and 99th percentiles. Store the result in `health_win`. Compare `summary(health_small$income)` to `summary(health_win$income)`.

```{r task2}
# YOUR CODE HERE
# p01 <- quantile(health_small$income, 0.01, na.rm = TRUE)
# p99 <- quantile(health_small$income, 0.99, na.rm = TRUE)
# health_win <- health_small |>
#   mutate(income = pmax(p01, pmin(p99, income)))
# summary(health_small$income)
# summary(health_win$income)
```

## Task 3 — Log transformation of claim_amount

For rows where `claim_amount > 0`, create side-by-side histograms of `claim_amount` (raw) and `log1p(claim_amount)`. Use base R `par(mfrow = c(1, 2))` and `hist()`.

```{r task3}
# YOUR CODE HERE
# pos_claims <- health_small$claim_amount[health_small$claim_amount > 0]
# par(mfrow = c(1, 2))
# hist(pos_claims,        main = "Claim Amount (raw)",    col = "steelblue")
# hist(log1p(pos_claims), main = "Claim Amount (log1p)", col = "darkorange")
# par(mfrow = c(1, 1))
```
