---
title: "BDAT 602 — Module 6, Lesson 1 Exercise"
subtitle: "Tokenisation and TF-IDF"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
source("R/simulate_bdat602_data.R")
health_small <- simulate_bdat602(n = 10000, seed = 602)
```

## Task 1 — Tokenise and count top words

Filter to rows where `complaint_notes` is not NA. Tokenise using `unnest_tokens()`, remove stop words with `anti_join(stop_words)`, and print the top 10 words by frequency.

```{r task1}
# YOUR CODE HERE
# library(tidytext); library(dplyr)
# data("stop_words")
# complaints <- health_small |>
#   filter(!is.na(complaint_notes)) |>
#   mutate(doc_id = row_number())
# tokens_clean <- complaints |>
#   unnest_tokens(word, complaint_notes) |>
#   anti_join(stop_words, by = "word") |>
#   filter(nchar(word) > 2)
# tokens_clean |> count(word, sort = TRUE) |> slice_head(n = 10)
```

## Task 2 — TF-IDF by plan tier

Compute TF-IDF treating `plan_tier` as the document. Show the top 3 discriminating words for each tier.

```{r task2}
# YOUR CODE HERE
# tokens_clean |>
#   count(plan_tier, word) |>
#   bind_tf_idf(term = word, document = plan_tier, n = n) |>
#   group_by(plan_tier) |>
#   slice_max(tf_idf, n = 3) |>
#   select(plan_tier, word, tf_idf) |>
#   print(n = 20)
```

## Task 3 — Document-Term Matrix

Build a DTM using `cast_dtm()` with `doc_id` as the document. Report its dimensions.

```{r task3}
# YOUR CODE HERE
# dtm <- tokens_clean |>
#   count(doc_id, word) |>
#   cast_dtm(document = doc_id, term = word, value = n)
# dim(dtm)
```
