---
title: "BDAT 602 — Module 6, Lesson 1 Solution"
subtitle: "Tokenisation and TF-IDF"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
source("R/simulate_bdat602_data.R")
health_small <- simulate_bdat602(n = 10000, seed = 602)
```

## Task 1 — Tokenise and count top words

```{r task1}
library(tidytext)
library(dplyr)

data("stop_words")

complaints <- health_small |>
  filter(!is.na(complaint_notes)) |>
  mutate(doc_id = row_number())

tokens_clean <- complaints |>
  unnest_tokens(word, complaint_notes) |>
  anti_join(stop_words, by = "word") |>
  filter(nchar(word) > 2)

tokens_clean |>
  count(word, sort = TRUE) |>
  slice_head(n = 10)
```

The most frequent words reflect the templated complaint categories: "claim", "premium", "coverage", "billing", "service".

## Task 2 — TF-IDF by plan tier

```{r task2}
tokens_clean |>
  count(plan_tier, word) |>
  bind_tf_idf(term = word, document = plan_tier, n = n) |>
  group_by(plan_tier) |>
  slice_max(tf_idf, n = 3) |>
  select(plan_tier, word, tf_idf)
```

Since the complaint templates are shared across plan tiers, TF-IDF discriminating words tend to be the less common complaint phrases that appear more in one tier than another.

## Task 3 — Document-Term Matrix

```{r task3}
dtm <- tokens_clean |>
  count(doc_id, word) |>
  cast_dtm(document = doc_id, term = word, value = n)

dim(dtm)
```

The DTM has ~9,000 rows (complaints) and several hundred unique terms after stop word removal. It is very sparse — most complaints use only a small subset of the vocabulary.
