---
title: "BDAT 608 | Module 4, Lesson 2 — SOLUTION"
author: "Instructor Reference"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 5)
```

```{r libraries}
library(tidyverse)
library(modelr)
options(na.action = na.warn)
data("sim1", package = "modelr")
data("sim3", package = "modelr")
sim1_base <- sim1
```

## Task 1 — Fit Models on sim3

```{r fit-sim3}
mod_add <- lm(y ~ x1 + x2,  data = sim3)
mod_int <- lm(y ~ x1 * x2,  data = sim3)

grid_sim3 <- sim3 |>
  data_grid(x1, x2) |>
  gather_predictions(mod_add, mod_int)

ggplot(sim3, aes(x1, y, colour = x2)) +
  geom_point(alpha = 0.6) +
  geom_line(data = grid_sim3, aes(y = pred), linewidth = 1) +
  facet_wrap(~ model, labeller = label_both) +
  labs(title = "Additive (+) vs Interaction (*) model on sim3") +
  theme_minimal(base_size = 12)
```

## Task 2 — Residual Comparison

```{r residuals-sim3}
sim3 |>
  gather_residuals(mod_add, mod_int) |>
  ggplot(aes(x1, resid, colour = x2)) +
  geom_point(alpha = 0.8) +
  geom_ref_line(h = 0) +
  facet_grid(model ~ x2) +
  scale_colour_brewer(palette = "Set1") +
  labs(title = "Residuals: additive vs interaction", y = "Residual") +
  theme_minimal(base_size = 11)
# mod_add residuals show systematic within-group patterns (slopes differ by group).
# mod_int residuals show random scatter — the interaction captures the slope differences.
```

## Task 3 — AIC Comparison

```{r aic-sim3}
AIC(mod_add, mod_int)
cat("Additive R²:   ", round(summary(mod_add)$r.squared, 4), "\n")
cat("Interaction R²:", round(summary(mod_int)$r.squared, 4), "\n")
# Interaction model has lower AIC → preferred for sim3.
```

## Task 4 — I() vs ^ in Formulas

```{r formula-demo}
# WRONG: y ~ x^2 + x collapses to y ~ x
model_matrix(sim1_base, y ~ x^2 + x) |> head(3)

# CORRECT: I() protects exponentiation
model_matrix(sim1_base, y ~ I(x^2) + x) |> head(3)
# Two distinct columns: (Intercept), I(x^2), x
```

## Reflection

If the interaction model has higher AIC than the additive model, it means the extra parameters (the `x1:x2` interaction terms) do not improve the fit enough to justify their complexity cost. The data do not show evidence that the slope of `x1` differs meaningfully across groups. This illustrates the **parsimony principle**: prefer the simpler model when the added complexity does not measurably improve predictive performance. AIC quantifies this trade-off by penalising each additional parameter.
