---
title: "BDAT 608 | Module 5, Lesson 2 — SOLUTION"
author: "Instructor Reference"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5)
```

```{r libraries}
library(tidyverse)
library(modelr)
library(MASS)
library(mgcv)
select <- dplyr::select; filter <- dplyr::filter
options(na.action = na.warn)
data("sim1", package = "modelr")
data("sim4", package = "modelr")
sim1_base <- sim1
```

## Task 1 — Inject Outliers and Fit Models

```{r robust-fit}
sim1_out <- sim1_base |>
  mutate(
    y          = ifelse(row_number() %in% c(3, 17), y + 15, y),
    is_outlier = row_number() %in% c(3, 17)
  )

mod_ols <- lm(y ~ x,                      data = sim1_out)
mod_rob <- MASS::rlm(y ~ x, method = "M", data = sim1_out)
```

## Task 2 — Plot Comparison

```{r robust-plot}
ggplot(sim1_out, aes(x, y, shape = is_outlier)) +
  geom_point(colour = "grey30", size = 2.5) +
  geom_abline(aes(intercept = coef(mod_ols)[1], slope = coef(mod_ols)[2],
                  colour = "OLS"),    linewidth = 1.2) +
  geom_abline(aes(intercept = coef(mod_rob)[1], slope = coef(mod_rob)[2],
                  colour = "Robust"), linewidth = 1.2) +
  scale_colour_manual(values = c("OLS" = "#C9A84C", "Robust" = "#1B3A6B")) +
  scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 8),
                     labels  = c("Normal", "Outlier"), name = "Point type") +
  labs(title = "OLS vs Robust regression with two outliers", colour = "Method") +
  theme_minimal(base_size = 13)
```

## Task 3 — Coefficient Comparison

```{r coef-compare}
cat("OLS:    intercept =", round(coef(mod_ols)[1], 2),
    " slope =", round(coef(mod_ols)[2], 2), "  <- pulled by outliers\n")
cat("Robust: intercept =", round(coef(mod_rob)[1], 2),
    " slope =", round(coef(mod_rob)[2], 2), "  <- close to true values (4.22, 2.05)\n")
```

## Task 4 — Fit GAM on sim4

```{r gam-fit}
mod_gam <- gam(y ~ s(x1) + s(x2), data = sim4, method = "REML")
summary(mod_gam)
```

## Task 5 — GAM Smooth Plots and Heatmap

```{r gam-plots}
plot(mod_gam, pages = 1, shade = TRUE, col = "#1B3A6B", shade.col = "#C9A84C40")
```

```{r gam-heatmap}
sim4 |>
  data_grid(x1 = seq_range(x1, n=40), x2 = seq_range(x2, n=40)) |>
  add_predictions(mod_gam) |>
  ggplot(aes(x1, x2, fill = pred)) +
  geom_tile() +
  scale_fill_viridis_c() +
  labs(title = "GAM predicted surface (additive smooths)", fill = "ŷ") +
  theme_minimal(base_size = 13)
```

## Reflection

edf = 1.02 for `s(x1)` means the estimated smooth is essentially a straight line — it has barely more complexity than a linear term. Yes, you could simplify: replace `s(x1)` with the linear term `x1` to reduce model complexity without meaningful loss of fit. A useful rule of thumb is: if edf < 1.5, the smooth is approximately linear and can be replaced with the corresponding fixed term.
