---
title: "BDAT 602 — Lecture 1"
subtitle: "What is Data Mining? The Big Data Landscape and the KDD Process"
date: "2026-05-18"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: show
    number_sections: true
---

```{r setup, include=FALSE}
# ── Pandoc (needed when knitting outside RStudio via Rscript) ────────────────
pandoc_path <- "/usr/lib/rstudio/resources/app/bin/quarto/bin/tools/x86_64"
if (nchar(Sys.getenv("RSTUDIO_PANDOC")) == 0 && dir.exists(pandoc_path)) {
  Sys.setenv(RSTUDIO_PANDOC = pandoc_path)
}

# ── Set working directory to project root for all chunks ────────────────────
# knitr defaults to the Rmd file's directory; this restores project-root paths
knitr::opts_knit$set(root.dir = normalizePath(".."))

# ── Clean slate ──────────────────────────────────────────────────────────────
rm(list = setdiff(ls(), c("knitr", "rmarkdown")))
gc()
options(scipen = 999, stringsAsFactors = FALSE)

# ── Global chunk options ─────────────────────────────────────────────────────
knitr::opts_chunk$set(
  echo       = TRUE,
  message    = FALSE,
  warning    = FALSE,
  fig.width  = 9,
  fig.height = 5,
  dpi        = 150
)
```

---

# What is Data Mining?

**Data mining** is the process of discovering patterns, anomalies, and actionable insights from large datasets using statistical, machine learning, and database techniques.

It is *not* just querying a database. Key distinctions:

| Task | Example |
|------|---------|
| **Classification** | Will this policyholder churn? (`churned`) |
| **Regression** | How much will this claim cost? (`claim_amount`) |
| **Clustering** | Which policyholders behave similarly? |
| **Association rules** | Which add-on covers are bought together? |
| **Anomaly detection** | Is this claim fraudulent? (`fraud_flag`) |
| **Text mining** | What are customers complaining about? (`complaint_notes`) |

All six tasks appear in the course dataset generated below.

---

# The Big Data Landscape — The 4 Vs

A dataset qualifies as "Big Data" when it stresses conventional tools across at least one of four dimensions:

| V | Definition | In our dataset |
|---|-----------|---------------|
| **Volume** | Scale of data | 500,000 records, 40 variables |
| **Velocity** | Speed of generation/processing | Claims arrive continuously |
| **Variety** | Heterogeneity of types | Numerics, categoricals, dates, free text |
| **Veracity** | Trustworthiness / quality | Intentional NAs, outliers, entry errors |

A fifth V — **Value** — is the goal: extract business decisions from the noise.

---

# The KDD Process

The **Knowledge Discovery in Databases (KDD)** process defines the end-to-end pipeline from raw data to deployed knowledge:

1. **Selection** — identify which data to use
2. **Preprocessing** — clean, handle missing values, remove noise
3. **Transformation** — encode, scale, reduce dimensions
4. **Data Mining** — apply the algorithm (cluster, classify, …)
5. **Interpretation / Evaluation** — validate and communicate results

> Data preparation (steps 1–3) consumes roughly **80 % of project time**. Lectures 2 onward address each step in depth.

The **CRISP-DM** framework operationalises KDD for industry projects, adding *Business Understanding* and *Deployment* phases and emphasising that the process is **cyclical** — evaluation findings loop back to earlier phases.

---

# Setup — Generate the Course Dataset

```{r load-data}
source("R/simulate_bdat602.R")

# 500,000-row dataset — matches the "Big Data" scale discussed above
health_data  <- simulate_bdat602(n = 500000, seed = 602)

# 10,000-row slice for interactive exploration
health_small <- simulate_bdat602(n = 10000,  seed = 602)
```

---

# A First Look at the Data

## Dimensions and variable names

```{r structure}
dim(health_data)
names(health_data)
```

The dataset has **`r nrow(health_data)`** rows and **`r ncol(health_data)`** columns across six groups: identifiers, demographics, policy details, medical usage, claims history, and customer behaviour.

## First three rows

```{r head}
head(health_data, 3)
```

## Full summary with skimr

`skimr::skim()` gives type-aware summaries (completion rate, percentiles, inline histograms) in one call — far more informative than `summary()` for mixed-type data.

```{r skim}
library(dplyr)
library(skimr)

skim(health_small)
```

## Plan tier distribution

```{r plan-tier}
health_small |>
  count(plan_tier) |>
  mutate(pct = round(100 * n / sum(n), 1)) |>
  arrange(desc(n))
```

## Claim amount distribution

```{r claim-summary}
health_small |>
  filter(claim_amount > 0) |>
  summarise(
    mean_claim   = round(mean(claim_amount),   2),
    median_claim = round(median(claim_amount), 2),
    sd_claim     = round(sd(claim_amount),     2),
    max_claim    = round(max(claim_amount),    2),
    pct_zero     = round(mean(health_small$claim_amount == 0) * 100, 1)
  )
```

> The mean is far above the median — a classic sign of right skew. `pct_zero` tells us how many policyholders made no claims at all. Both facts will drive our preprocessing choices in Lecture 2.

## Fraud prevalence

```{r fraud-rate}
health_small |>
  summarise(fraud_rate_pct = round(mean(fraud_flag) * 100, 2))
```

About 3 % of records are flagged as fraudulent — a **rare event**. A naïve model that predicts "no fraud" for every record achieves 97 % accuracy but zero usefulness. Class imbalance remedies are covered in Lecture 2.

---

# Loading Big Data into Spark {.tabset}

For 500,000+ rows, R's in-memory operations slow down. `sparklyr` lets you run `dplyr` verbs on a distributed Spark cluster. Computation stays inside Spark; only final summaries are `collect()`ed back to R.

> The chunks below require a running Spark session — shown for reference (`eval=FALSE`). Run them interactively when Spark is available.

## Connect and copy

```{r spark-connect, eval=FALSE}
library(sparklyr)
library(dplyr)

sc <- spark_connect(master = "local[*]")

# Copy the in-memory data frame to Spark
# From this point, health_tbl lives in Spark, not in R RAM
health_tbl <- copy_to(
  sc,
  health_data,
  name      = "health_insurance",
  overwrite = TRUE
)

health_tbl |> count()     # count rows inside Spark
health_tbl |> glimpse()   # inspect schema
```

## Sample for prototyping

```{r spark-sample, eval=FALSE}
# 10 % Bernoulli sample — keeps work fast during development
health_sample <- health_tbl |>
  sdf_sample(fraction = 0.10, replacement = FALSE, seed = 602)

sdf_nrow(health_sample)   # ~50,000 rows
```

> **Key rule:** `filter()`, `mutate()`, and `group_by()` are translated to Spark SQL and run distributed. Only `collect()` brings results back to R. Never `collect()` the full 500 k rows — only collect summaries.

---

# Reading and Inspecting Data from a CSV {.tabset}

In practice you will read data from flat files rather than generating it. Two main options depending on file size:

| Package | Function | Best for |
|---------|---------|---------|
| `readr` | `read_csv()` | Tidyverse workflows, moderate files |
| `data.table` | `fread()` | Maximum speed on large files |

```{r save-csv, include=FALSE}
# Write health_small to CSV so the read examples below work
library(readr)
write_csv(health_small, "data/raw/health_insurance_small.csv")
```

## readr

```{r read-readr}
library(readr)

df_readr <- read_csv("data/raw/health_insurance_small.csv")
glimpse(df_readr)
```

## data.table

```{r read-datatable}
library(data.table)

df_dt <- fread("data/raw/health_insurance_small.csv")
cat("Rows:", nrow(df_dt), "| Cols:", ncol(df_dt), "\n")
cat("Class:", class(df_dt)[1], "\n")
```

## Quick inspection

```{r inspect}
# Missing values per column
colSums(is.na(df_readr)) |> sort(decreasing = TRUE) |> (\(x) x[x > 0])()

# Frequency table for a nominal variable
df_readr |>
  count(region, sort = TRUE) |>
  mutate(pct = round(100 * n / sum(n), 1))

# Numeric summary for key variables
df_readr |>
  select(age, bmi, income, claim_amount) |>
  summary()
```

---

# First Exploratory Visualisations

## Distribution of claim amounts

Claim amounts are heavily right-skewed — many small claims and a long tail of large ones. A log scale reveals the shape of the bulk distribution.

```{r plot-claims}
library(ggplot2)

ggplot(health_small |> filter(claim_amount > 0),
       aes(x = claim_amount)) +
  geom_histogram(bins = 60, fill = "steelblue", colour = "white") +
  scale_x_log10(labels = scales::comma) +
  labs(title = "Distribution of Claim Amounts (log scale)",
       x = "Claim Amount (USD, log scale)", y = "Count") +
  theme_minimal()
```

## Claim amount by plan tier

```{r plot-tier}
ggplot(health_small |> filter(claim_amount > 0),
       aes(x = plan_tier, y = claim_amount, fill = plan_tier)) +
  geom_boxplot(outlier.alpha = 0.2) +
  scale_y_log10(labels = scales::comma) +
  labs(title = "Claim Amount by Plan Tier",
       x = "Plan Tier", y = "Claim Amount (log scale)") +
  theme_minimal() +
  theme(legend.position = "none")
```

Higher plan tiers show higher median claims — both because Platinum attracts higher-risk enrollees and because higher coverage incentivises larger claims.

---

# Automated EDA with DataExplorer

`DataExplorer::create_report()` generates a full HTML EDA report in one line — missing value profiles, distributions, correlations, and QQ plots. Essential for the first pass on any new dataset.

## Missing value map

```{r eda-missing}
library(DataExplorer)

plot_missing(health_small)
```

## Continuous variable distributions

```{r eda-hist, fig.height=8}
plot_histogram(health_small)
```

## Correlation heatmap

```{r eda-corr}
plot_correlation(
  health_small |>
    select(age, bmi, income, premium,
           claim_amount, num_claims,
           num_chronic_conditions),
  title = "Correlation Matrix: Key Numeric Variables"
)
```

## Full automated report

```{r eda-report, eval=FALSE}
# Generates output/reports/lecture1_eda.html — open in a browser
create_report(
  health_small,
  output_file  = "lecture1_eda.html",
  output_dir   = "output/reports",
  report_title = "BDAT 602: Health Insurance EDA"
)
```

---

# Key Takeaways

* **Data mining** covers classification, regression, clustering, association rules, anomaly detection, and text mining — all demonstrated on the course dataset.
* The **4 Vs** (Volume, Velocity, Variety, Veracity) characterise Big Data; the course dataset is designed to exercise all four.
* The **KDD process** places data preparation before any algorithm. Expect to spend 80 % of your time here.
* `skimr::skim()` and `DataExplorer::create_report()` are your first-pass EDA tools.
* `sparklyr` lets you run `dplyr` verbs at scale — push computation to Spark, `collect()` only summaries.

**Looking ahead — Lecture 2:** Data preparation in depth: handling missing values (MCAR / MAR / MNAR), outlier detection and treatment, scaling, categorical encoding, PCA, and building a leak-proof `recipes` preprocessing pipeline.

---

# Session Info

```{r session-info}
sessionInfo()
```
