---
title: "BDAT 602 — Lecture 2"
subtitle: "Data Preparation: Cleaning, Transforming, and Exploring Data"
date: "2026-05-18"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: show
    number_sections: true
---

```{r setup, include=FALSE}
# ── Pandoc (needed when knitting outside RStudio via Rscript) ────────────────
pandoc_path <- "/usr/lib/rstudio/resources/app/bin/quarto/bin/tools/x86_64"
if (nchar(Sys.getenv("RSTUDIO_PANDOC")) == 0 && dir.exists(pandoc_path)) {
  Sys.setenv(RSTUDIO_PANDOC = pandoc_path)
}

# ── Set working directory to project root for all chunks ────────────────────
knitr::opts_knit$set(root.dir = normalizePath(".."))

# ── Clean slate ──────────────────────────────────────────────────────────────
rm(list = setdiff(ls(), c("knitr", "rmarkdown")))   # keep knitr objects
gc()
options(scipen = 999, stringsAsFactors = FALSE)

# ── Global chunk options ──────────────────────────────────────────────────────
knitr::opts_chunk$set(
  echo       = TRUE,
  message    = FALSE,
  warning    = FALSE,
  fig.width  = 9,
  fig.height = 5,
  dpi        = 150
)
```

---

# Setup

Source the data generator and shared helpers, then create the two working datasets used throughout this lecture.

```{r load-data}
source("R/simulate_bdat602.R")
source("R/utils.R")

# 500 k rows — used for Spark and the final pipeline
health_data  <- simulate_bdat602(n = 500000, seed = 602)

# 10 k rows — used for interactive R sections (faster knitting)
health_small <- simulate_bdat602(n = 10000,  seed = 602)

library(dplyr)
library(skimr)
```

A quick structural check before anything else:

```{r structure}
dim(health_data)
```

The dataset has **500,000 rows** and **40 columns** covering demographics, policy details, claims, engagement, and three classification targets (`fraud_flag`, `churned`, `high_risk`).

---

# Diagnosing Missing Values

Before deciding *how* to handle missing data you must first understand *why* it is missing. The mechanism determines the remedy:

| Mechanism | Definition | Safe remedy |
|-----------|-----------|-------------|
| **MCAR** — Missing Completely At Random | Missingness is unrelated to any variable | Deletion or imputation both fine |
| **MAR** — Missing At Random | Missingness depends on *observed* variables | Imputation (kNN, model-based) |
| **MNAR** — Missing Not At Random | Missingness depends on the *missing value itself* | Flag + domain knowledge |

## Count and rank NAs

```{r missing-count}
# Columns with at least one NA, sorted by count
colSums(is.na(health_small)) |>
  sort(decreasing = TRUE) |>
  (\(x) x[x > 0])()
```

```{r missing-pct}
# As a percentage
round(colMeans(is.na(health_small)) * 100, 2) |>
  sort(decreasing = TRUE) |>
  (\(x) x[x > 0])()
```

## Visual missing-value map

```{r missing-plot}
library(DataExplorer)
plot_missing(health_small)
```

> **Key observations:** `bmi` (~5 %) and `income` (~4 %) are likely **MAR** — patients who do not visit a GP are less likely to have a recorded BMI. `complaint_notes` is **MNAR** — only policyholders who lodged a complaint have a value.

---

# Simple Imputation — Median, Mode, and Flags

Simple imputation is fast and transparent. Always add a **binary missingness indicator** alongside any imputed numeric variable so downstream models can learn from the pattern of missingness itself.

```{r simple-imputation}
health_clean <- health_small

# ── Strategy 1: Numeric → median (robust to skew) ──────────────────────────
bmi_median    <- median(health_clean$bmi,    na.rm = TRUE)
income_median <- median(health_clean$income, na.rm = TRUE)

health_clean <- health_clean |>
  mutate(
    bmi_missing    = as.integer(is.na(bmi)),
    income_missing = as.integer(is.na(income)),
    bmi            = if_else(is.na(bmi),    bmi_median,    bmi),
    income         = if_else(is.na(income), income_median, income)
  )

# ── Strategy 2: Categorical → mode (from R/utils.R) ────────────────────────
health_clean <- health_clean |>
  mutate(
    employment_type = if_else(
      is.na(employment_type),
      get_mode(employment_type),
      employment_type
    )
  )

# ── Strategy 3: Free-text → treat NA as "no complaint" ─────────────────────
health_clean <- health_clean |>
  mutate(
    has_complaint   = as.integer(!is.na(complaint_notes)),
    complaint_notes = if_else(is.na(complaint_notes),
                               "no complaint", complaint_notes)
  )

# Verify
colSums(is.na(health_clean[, c("bmi", "income", "complaint_notes")]))
```

---

# kNN Imputation for Better Accuracy

Median imputation replaces every missing value with the *same* number regardless of context. **kNN imputation** instead finds the *k* most similar complete records and averages their values — far more accurate when data is MAR.

The critical rule: **always use `recipes`** so the *k* neighbours are identified from training data only. Fitting on the full dataset (including what will become the test set) is data leakage.

```{r knn-imputation}
library(recipes)

impute_recipe <- recipe(claim_amount ~ ., data = health_small) |>
  step_impute_knn(
    bmi,
    neighbors   = 5,
    impute_with = imp_vars(age, sex, region, employment_type, plan_tier)
  ) |>
  step_impute_median(income)   # income is MNAR → median is safer

impute_prep    <- prep(impute_recipe, training = health_small)
health_imputed <- bake(impute_prep, new_data = health_small)

summary(health_imputed$bmi)
```

> **Why `recipes`?** The median and the *k* neighbours are learned during `prep()` from training data only, then applied identically to any new data via `bake()`. This prevents leakage and ensures reproducibility.

---

# Handling Missing Values in Spark at Scale {.tabset}

For 500,000+ rows the in-memory approaches above are impractical. `sparklyr` lets you count NAs and compute medians **inside Spark** without ever moving the raw data into R.

> **Note:** The chunks below require a running Spark session. They are shown for reference (`eval=FALSE`). Run them interactively when Spark is available.

## Count NAs in Spark

```{r spark-missing, eval=FALSE}
library(sparklyr)
library(dplyr)

sc         <- spark_connect(master = "local[*]")
health_tbl <- copy_to(sc, health_data,
                       name = "health_insurance", overwrite = TRUE)

health_tbl |>
  summarise(
    bmi_missing    = sum(as.integer(is.na(bmi)),    na.rm = TRUE),
    income_missing = sum(as.integer(is.na(income)), na.rm = TRUE)
  )
```

## Impute median in Spark

```{r spark-impute, eval=FALSE}
# percentile_approx() is Spark's fast approximate median
bmi_med_spark <- health_tbl |>
  summarise(med = percentile_approx(bmi, 0.5, 1000L)) |>
  collect() |>
  pull(med)

health_tbl_clean <- health_tbl |>
  mutate(
    bmi_missing = as.integer(is.na(bmi)),
    bmi         = if_else(is.na(bmi), bmi_med_spark, bmi)
  )

sdf_nrow(health_tbl_clean)   # confirm no rows dropped
```

---

# Outlier Detection

An **outlier** is an observation that deviates markedly from the general pattern. It may be a genuine extreme value, a measurement error, or — crucially for fraud detection — **the signal itself**.

Three types to recognise:

* **Global**: extreme relative to the whole dataset (BMI = 200)
* **Contextual**: extreme only within a context (£50k claim on a Bronze plan)
* **Collective**: a group that is anomalous together (weekend claims from the same agent)

## Visual — box plot

```{r outlier-boxplot}
library(ggplot2)

ggplot(health_clean, aes(y = bmi)) +
  geom_boxplot(fill = "steelblue", outlier.colour = "red",
               outlier.alpha = 0.4) +
  labs(title = "BMI Distribution with Outliers", y = "BMI") +
  theme_minimal()
```

## Statistical — IQR rule

Values below $Q_1 - 1.5 \times \text{IQR}$ or above $Q_3 + 1.5 \times \text{IQR}$ are flagged.

```{r outlier-iqr}
q1_bmi  <- quantile(health_clean$bmi, 0.25, na.rm = TRUE)
q3_bmi  <- quantile(health_clean$bmi, 0.75, na.rm = TRUE)
iqr_bmi <- q3_bmi - q1_bmi

health_clean <- health_clean |>
  mutate(
    bmi_outlier = bmi < (q1_bmi - 1.5 * iqr_bmi) |
                  bmi > (q3_bmi + 1.5 * iqr_bmi)
  )

health_clean |> count(bmi_outlier)
```

## Statistical — Z-score

The Z-score $|z_i| = \left|\frac{x_i - \bar{x}}{s}\right| > 3$ assumes approximate normality. Use on roughly symmetric variables like income (after any transformation).

```{r outlier-zscore}
health_clean <- health_clean |>
  mutate(
    income_z       = (income - mean(income, na.rm = TRUE)) /
                      sd(income, na.rm = TRUE),
    income_outlier = abs(income_z) > 3
  )

health_clean |> count(income_outlier)
```

---

# Treating Outliers

Detection alone is not enough — you must decide what to do. The right action depends on *why* the outlier exists.

| Treatment | When to use |
|-----------|------------|
| **Remove** | Clear data-entry error |
| **Winsorise** (cap) | Genuine extreme; want to retain the record |
| **Flag + retain** | Outlier is informative (fraud detection) |
| **Log-transform** | Right-skewed variable; all values valid |

## Winsorisation

```{r winsorise}
p01_income <- quantile(health_clean$income, 0.01, na.rm = TRUE)
p99_income <- quantile(health_clean$income, 0.99, na.rm = TRUE)

health_clean <- health_clean |>
  mutate(
    bmi    = pmax(10, pmin(60, bmi)),           # biologically plausible range
    income = pmax(p01_income, pmin(p99_income, income))
  )

health_clean |> select(bmi, income) |> summary()
```

## Log transformation

Right-skewed variables like income compress poorly with standardisation. Taking $\log(1 + x)$ makes the distribution more symmetric, which benefits distance-based algorithms (k-means, kNN) and linear models.

```{r log-transform, fig.height=4}
health_clean <- health_clean |>
  mutate(log_income = log1p(income))

par(mfrow = c(1, 2))
hist(health_clean$income,     main = "Income (raw)",
     col = "steelblue", border = "white", xlab = "")
hist(health_clean$log_income, main = "Income (log1p)",
     col = "darkorange", border = "white", xlab = "")
par(mfrow = c(1, 1))
```

---

# Scaling Numeric Variables

**Why it matters:** compute the distance between two policyholders using age (range ~70) and income (range ~$500,000). Income completely dominates — age becomes irrelevant.

Algorithms that require scaling: **k-means**, **kNN**, **PCA**, hierarchical clustering, SVM, ridge/lasso regression.
Algorithms unaffected: decision trees, random forests, naïve Bayes, association rules.

| Method | Formula | Range | Robust to outliers? |
|--------|---------|-------|-------------------|
| **Z-score** | $(x - \bar{x}) / s$ | $(-\infty, +\infty)$ | No |
| **Min-Max** | $(x - x_{\min}) / (x_{\max} - x_{\min})$ | $[0, 1]$ | No |
| **Robust** | $(x - \text{median}) / \text{IQR}$ | $(-\infty, +\infty)$ | Yes |

## Z-score via recipes

```{r scale-zscore}
library(recipes)

num_vars <- c("age", "bmi", "income", "premium", "deductible",
              "coverage_amount", "policy_age_months",
              "num_chronic_conditions", "num_visits",
              "num_prescriptions", "num_hospital_admissions",
              "num_claims", "avg_past_claim", "claim_amount",
              "app_logins_monthly", "support_calls")

scale_recipe <- recipe(~ ., data = health_clean) |>
  step_normalize(all_of(num_vars))

scale_prep    <- prep(scale_recipe, training = health_clean)
health_scaled <- bake(scale_prep, new_data = health_clean)

# Verify: mean ~0, sd ~1
health_scaled |>
  select(age, bmi, income) |>
  summarise(across(everything(), list(mean = mean, sd = sd),
                   .names = "{.col}_{.fn}"))
```

## Min-Max (manual)

```{r scale-minmax}
# min_max_scale() lives in R/utils.R
health_clean <- health_clean |>
  mutate(
    age_mm = min_max_scale(age),
    bmi_mm = min_max_scale(bmi)
  )

summary(health_clean[, c("age_mm", "bmi_mm")])
```

---

# Encoding Categorical Variables

Most mining algorithms require numeric input. The naive approach — assigning integers 1, 2, 3, … — imposes a false ordering. Use the right encoding for the variable type.

| Variable type | Encoding | `recipes` step |
|--------------|---------|---------------|
| Nominal (sex, region) | One-hot / dummy (k-1 columns) | `step_dummy()` |
| Ordinal (education, plan_tier) | Integer preserving order | `step_integer()` |
| High-cardinality nominal | Target encoding | `step_lencode_*()` |

> **The dummy trap:** always drop one category per variable (the reference level). `recipes` does this automatically with `one_hot = FALSE`.

```{r encoding}
library(recipes)

encode_recipe <- recipe(claim_amount ~ ., data = health_clean) |>
  # Nominal: k-1 dummy columns
  step_dummy(sex, region, payment_method,
             employment_type, one_hot = FALSE) |>
  # Ordinal: set explicit level ordering, then integer-encode
  step_mutate(
    education = factor(education,
                       levels = c("None", "Primary", "Secondary",
                                   "Tertiary", "Postgraduate"),
                       ordered = TRUE),
    plan_tier = factor(plan_tier,
                       levels = c("Bronze", "Silver", "Gold", "Platinum"),
                       ordered = TRUE)
  ) |>
  step_integer(education, plan_tier) |>
  step_nzv(all_predictors())   # drop near-zero variance artefacts

encode_prep    <- prep(encode_recipe, training = health_clean)
health_encoded <- bake(encode_prep, new_data = health_clean)

# First 30 column names after encoding
names(health_encoded) |> head(30)
```

```{r encoding-verify}
# Confirm no character columns remain
sapply(health_encoded, class) |> table()
```

---

# Principal Component Analysis (PCA)

PCA addresses the **curse of dimensionality**: as the number of variables grows, the feature space expands exponentially and distance-based algorithms degrade. PCA creates a smaller set of new variables (principal components) that capture most of the variance.

* **PC1** — direction of greatest variance
* **PC2** — direction of second-greatest variance, perpendicular to PC1
* Each subsequent PC captures less variance
* PCs are **uncorrelated** — no multicollinearity

> **In BDAT 602** we use PCA as a preprocessing step before clustering and as a visualisation tool. The full mathematical treatment is in BDAT 604.

**Requirements:** numeric, scaled input only — cannot be applied to raw categorical variables.

## Fit PCA

```{r pca-fit}
library(broom)

pca_vars <- c("age", "bmi", "income", "premium", "deductible",
              "num_chronic_conditions", "num_visits",
              "num_prescriptions", "num_hospital_admissions",
              "num_claims", "avg_past_claim", "claim_amount",
              "app_logins_monthly", "support_calls",
              "policy_age_months")

pca_recipe <- recipe(~ ., data = health_clean[, pca_vars]) |>
  step_normalize(all_numeric_predictors()) |>
  step_pca(all_numeric_predictors(), num_comp = 5)

pca_prep   <- prep(pca_recipe)
pca_scores <- bake(pca_prep, new_data = NULL)

head(pca_scores)
```

## Scree plot

The **scree plot** shows how much variance each PC explains. Look for the "elbow" — the point where adding more PCs yields diminishing returns. Keep enough PCs to explain ≥ 80 % of total variance (or apply the Kaiser rule: eigenvalue > 1).

```{r pca-scree}
library(ggplot2)

tidy(pca_prep, number = 2, type = "variance") |>
  filter(terms == "percent variance") |>
  ggplot(aes(x = component, y = value)) +
  geom_col(fill = "steelblue") +
  geom_line(group = 1, colour = "red") +
  labs(title = "Scree Plot: Variance Explained by Each PC",
       x = "Principal Component", y = "% Variance Explained") +
  theme_minimal()
```

## PC loadings

Loadings show how much each original variable contributes to each component. Positive and negative bars indicate the direction of contribution.

```{r pca-loadings, fig.height=6}
tidy(pca_prep, number = 2, type = "coef") |>
  filter(component %in% c("PC1", "PC2")) |>
  ggplot(aes(x = reorder(terms, value), y = value,
             fill = value > 0)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ component) +
  coord_flip() +
  scale_fill_manual(values = c("steelblue", "darkorange")) +
  labs(title = "PCA Loadings: PC1 and PC2",
       x = NULL, y = "Loading") +
  theme_minimal()
```

## Score plot

Plot PC1 vs PC2 coloured by plan tier. If groups separate along a component, that component captures a risk/value dimension.

```{r pca-scoreplot}
pca_scores |>
  bind_cols(health_clean |> select(plan_tier)) |>
  ggplot(aes(x = PC1, y = PC2, colour = plan_tier)) +
  geom_point(alpha = 0.15, size = 0.8) +
  labs(title = "PCA Score Plot: PC1 vs PC2 by Plan Tier",
       colour = "Plan Tier") +
  theme_minimal()
```

---

# Exploratory Data Analysis (EDA)

EDA is **not** just plotting. It is a systematic investigation of distribution shapes, central tendency, missing values, outliers, relationships between variables, and class balance. The decisions made during EDA — what to impute, what to transform, what to encode — directly determine the quality of mining results.

## Univariate EDA

```{r eda-univariate, fig.width=11}
library(patchwork)

p1 <- health_clean |>
  filter(claim_amount > 0) |>
  ggplot(aes(x = claim_amount)) +
  geom_histogram(bins = 60, fill = "steelblue", colour = "white") +
  scale_x_log10(labels = scales::comma) +
  labs(title = "Claim Amount (log scale)", x = "USD", y = "Count") +
  theme_minimal()

p2 <- health_clean |>
  count(plan_tier) |>
  ggplot(aes(x = reorder(plan_tier, -n), y = n, fill = plan_tier)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = scales::comma(n)), vjust = -0.3, size = 3) +
  labs(title = "Policies by Plan Tier", x = NULL, y = "Count") +
  theme_minimal()

p1 + p2
```

## Bivariate EDA

```{r eda-bivariate}
health_clean |>
  filter(claim_amount > 0) |>
  ggplot(aes(x = plan_tier, y = claim_amount, fill = plan_tier)) +
  geom_boxplot(outlier.alpha = 0.15) +
  scale_y_log10(labels = scales::comma) +
  labs(title = "Claim Amount by Plan Tier",
       x = NULL, y = "Claim Amount (log scale)") +
  theme_minimal() +
  theme(legend.position = "none")
```

## Correlation heatmap

```{r eda-correlation, fig.height=6}
health_clean |>
  select(age, bmi, income, premium, deductible,
         num_chronic_conditions, num_visits, num_claims,
         claim_amount, support_calls, customer_rating) |>
  plot_correlation(type  = "continuous",
                   title = "Correlation Matrix: Numeric Variables")
```

## Fraud rate by employment type

```{r eda-fraud}
health_clean |>
  group_by(employment_type) |>
  summarise(fraud_rate = round(mean(fraud_flag) * 100, 2),
            n          = n()) |>
  arrange(desc(fraud_rate))
```

---

# EDA at Scale with sparklyr {.tabset}

With 500,000 rows, R-side `group_by()` operations can be slow. Push aggregations to Spark and `collect()` only the summary.

> The chunks below require a running Spark session — set `eval=FALSE` for knitting without Spark.

## Global summaries

```{r spark-eda-global, eval=FALSE}
library(sparklyr)
library(dplyr)

# sc and health_tbl created in Section 4
health_tbl |>
  summarise(
    mean_age       = mean(age,          na.rm = TRUE),
    mean_bmi       = mean(bmi,          na.rm = TRUE),
    mean_claim     = mean(claim_amount, na.rm = TRUE),
    median_claim   = percentile_approx(claim_amount, 0.5),
    fraud_rate_pct = mean(fraud_flag,   na.rm = TRUE) * 100,
    churn_rate_pct = mean(churned,      na.rm = TRUE) * 100
  ) |>
  collect()
```

## Group-wise summaries

```{r spark-eda-group, eval=FALSE}
health_tbl |>
  group_by(plan_tier) |>
  summarise(
    n           = n(),
    avg_premium = mean(premium,      na.rm = TRUE),
    avg_claim   = mean(claim_amount, na.rm = TRUE),
    pct_fraud   = mean(fraud_flag,   na.rm = TRUE) * 100,
    pct_churned = mean(churned,      na.rm = TRUE) * 100
  ) |>
  arrange(desc(avg_claim)) |>
  collect()
```

## Practice exercise

```{r spark-exercise, eval=FALSE}
# Fraud rate by region and employment type
health_tbl |>
  group_by(region, employment_type) |>
  summarise(fraud_rate = mean(fraud_flag, na.rm = TRUE) * 100,
            n          = n()) |>
  arrange(desc(fraud_rate)) |>
  collect()
```

---

# Class Imbalance

A critical EDA finding for any classification problem. When a target has very unequal class frequencies, standard algorithms default to predicting the majority class.

```{r class-imbalance}
# fraud_flag: ~3 % positive — a model predicting "no fraud" for everyone
# achieves 97 % accuracy but is completely useless
health_clean |>
  summarise(
    fraud_rate_pct  = round(mean(fraud_flag)                * 100, 2),
    churn_rate_pct  = round(mean(churned, na.rm = TRUE)      * 100, 2),
    high_risk_pct   = round(mean(high_risk, na.rm = TRUE)    * 100, 2)
  )
```

Common remedies (applied *after* the train/test split, on training data only):

1. **SMOTE** — generate synthetic minority records
2. **Random oversampling** — duplicate minority records
3. **Random undersampling** — remove majority records
4. **Class weights** — penalise misclassifying the minority more heavily
5. **Threshold adjustment** — lower the classification threshold from 0.5

---

# Complete Preprocessing Pipeline

Ad-hoc preprocessing causes **data leakage** (imputing with the whole-dataset mean before splitting), is hard to reproduce, and cannot be applied consistently to new data. A `recipes` pipeline solves all three problems.

The workflow:

```
recipe()   → define steps (blueprint only)
prep()     → learn parameters from training data
bake()     → apply to training set, test set, or new data
```

```{r pipeline-split}
library(tidymodels)

set.seed(602)
data_split <- initial_split(health_clean, prop = 0.75,
                             strata = plan_tier)
train_data <- training(data_split)
test_data  <- testing(data_split)

cat("Train rows:", nrow(train_data), "\n")
cat("Test rows: ", nrow(test_data),  "\n")
```

```{r pipeline-recipe}
prep_recipe <- recipe(claim_amount ~ ., data = train_data) |>
  # 1. Imputation
  step_impute_median(bmi, income) |>
  step_mutate(
    income_missing = as.integer(is.na(income)),
    bmi_missing    = as.integer(is.na(bmi))
  ) |>
  # 2. Outlier treatment
  step_mutate(
    bmi    = pmax(10, pmin(60, bmi)),
    income = pmax(quantile(income, 0.01, na.rm = TRUE),
                  pmin(quantile(income, 0.99, na.rm = TRUE), income))
  ) |>
  # 3. Log-transform skewed variables
  step_log(income, claim_amount, base = 10, offset = 1) |>
  # 4. Encode categoricals
  step_dummy(sex, region, payment_method, employment_type) |>
  step_mutate(
    education = factor(education,
                       levels = c("None", "Primary", "Secondary",
                                   "Tertiary", "Postgraduate"),
                       ordered = TRUE),
    plan_tier = factor(plan_tier,
                       levels = c("Bronze", "Silver", "Gold", "Platinum"),
                       ordered = TRUE)
  ) |>
  step_integer(education, plan_tier) |>
  # 5. Scale
  step_normalize(all_numeric_predictors()) |>
  # 6. Dimensionality reduction
  step_nzv(all_predictors()) |>
  step_corr(all_numeric_predictors(), threshold = 0.95)
```

```{r pipeline-prep-bake}
prep_fit        <- prep(prep_recipe, training = train_data)
train_processed <- bake(prep_fit, new_data = NULL)
test_processed  <- bake(prep_fit, new_data = test_data)

cat("Columns after pipeline — train:", ncol(train_processed),
    "| test:", ncol(test_processed), "\n")
cat("Missing values remaining:", sum(colSums(is.na(train_processed))), "\n")
```

```{r pipeline-verify}
# Which columns survived nzv + corr removal?
cat("Columns kept:", ncol(train_processed), "\n")
cat("Removed columns include those with near-zero variance or r > 0.95\n\n")

# Means ~0 and SDs ~1 — pick the first 4 numeric columns that survived
train_processed |>
  select(where(is.numeric)) |>
  select(1:4) |>
  summarise(across(everything(), list(mean = mean, sd = sd),
                   .names = "{.col}_{.fn}"))
```

```{r pipeline-save}
saveRDS(train_processed, "data/processed/train_processed.rds")
saveRDS(test_processed,  "data/processed/test_processed.rds")
cat("Processed datasets saved to data/processed/\n")
```

---

# Key Takeaways

| Problem | Remedy | `recipes` step |
|---------|--------|---------------|
| Missing values | Median / kNN imputation + flag | `step_impute_*` |
| Outliers | Winsorise or log-transform | `step_mutate`, `step_log` |
| Different scales | Z-score or Min-Max | `step_normalize` |
| Categorical variables | One-hot or ordinal encoding | `step_dummy`, `step_integer` |
| Too many variables | PCA | `step_pca` |

**The three rules of data preparation:**

1. Understand *why* values are missing before deciding how to handle them.
2. Always build a pipeline — never transform data by hand.
3. EDA is not optional and not a one-time step.

---

# Session Info

```{r session-info}
sessionInfo()
```
