---
title: "BDAT 602 — Module 2, Lesson 1 Exercise"
subtitle: "Diagnosing and Handling Missing Values"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
source("R/simulate_bdat602_data.R")
health_small <- simulate_bdat602(n = 10000, seed = 602)
```

## Task 1 — Count NAs

Count the number of missing values in `bmi`, `income`, and `complaint_notes`.

```{r task1}
# YOUR CODE HERE
# colSums(is.na(health_small[, c("bmi", "income", "complaint_notes")]))
```

## Task 2 — Median imputation with indicator

Impute `bmi` with its median. Add a `bmi_missing` indicator column (1 if originally missing, 0 otherwise). Store the result in `health_clean`. Verify that `bmi` has zero NAs after imputation.

```{r task2}
# YOUR CODE HERE
# library(dplyr)
# bmi_median <- median(health_small$bmi, na.rm = TRUE)
# health_clean <- health_small |>
#   mutate(
#     bmi_missing = as.integer(is.na(bmi)),
#     bmi         = if_else(is.na(bmi), bmi_median, bmi)
#   )
# colSums(is.na(health_clean[, c("bmi", "bmi_missing")]))
```

## Task 3 — recipes imputation

Build a `recipes` pipeline that applies `step_impute_median()` to `income`. Call `prep()` and `bake()` on `health_small`. Confirm `income` has zero NAs in the baked result.

```{r task3}
# YOUR CODE HERE
# library(recipes)
# impute_rec  <- recipe(claim_amount ~ ., data = health_small) |>
#   step_impute_median(income)
# impute_prep <- prep(impute_rec, training = health_small)
# health_imp  <- bake(impute_prep, new_data = health_small)
# sum(is.na(health_imp$income))
```
